---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ggblanket

<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/ggblanket)](https://CRAN.R-project.org/package=ggblanket)
<!-- badges: end -->

The goal of ggblanket is to assist users make beautiful ggplots faster and with less brainpower - while building intuition for how ggplot2 works.

## Installation

You can install the development version of ggblanket from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("davidhodge931/ggblanket")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(ggblanket)
## basic example code
```


```{r cars}
summary(cars)
```

## Wrapper functions available

* `gg_area`	
* `gg_bar`	
* `gg_blank`	
* `gg_boxplot`	
* `gg_col`	
* `gg_crossbar`
* `gg_density`
* `gg_errorbar`	
* `gg_freqpoly`	
* `gg_function`	
* `gg_histogram`	
* `gg_jitter`	
* `gg_label`
* `gg_line`
* `gg_linerange`
* `gg_path`	
* `gg_point`	
* `gg_pointrange`	
* `gg_polygon`
* `gg_qq`
* `gg_raster`	
* `gg_rect`
* `gg_ribbon`
* `gg_rug`
* `gg_segment`	
* `gg_sf`	
* `gg_smooth`	
* `gg_step`	
* `gg_text`	
* `gg_theme`
* `gg_tile`	
* `gg_violin`	
