---
title: "ggblanket"
author: "David Hodge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggblanket}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.asp = 0.6,
  out.width = "70%",
  dpi = 300)
```

## Overview

ggblanket is a package of ggplot2 wrapper functions.

The primary objective is to **simplify ggplot2 visualisation**.

Secondary objectives relate to:

* Design: produce well-designed visualisation
* Alignment: align with ggplot2 and tidyverse
* Scope: cover much of what ggplot2 does.

Computational speed has been traded-off.

## How it works


1.  First setup with `set_blanket()`
2.  Each `gg_*` function wraps a geom
3.  A merged `col` argument to colour/fill by a variable
4.  A `facet` argument to facet by a variable
5.  A `facet2` argument to facet by a 2nd variable
6.  Other aesthetics via `mapping` argument
7.  Prefixed arguments to customise x/y/col/facet
8.  Smart `*_label` defaults for axis and legend titles
9.  Other `ggplot2::geom_*` arguments via `...`
10. A 'symmetric' continuous scale with easy opt-out
11. Ability to add multiple `geom_*` layers
12. Families of `*_mode_*` themes with legend variants
13. Arguments to customise setup with `set_blanket()`
14. A `gg_blanket()` function with `geom` flexibility

```{r setup}
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)
library(ggblanket)
library(palmerpenguins)
library(patchwork)

penguins2 <- penguins |> 
  labelled::set_variable_labels(
    bill_length_mm = "Bill length (mm)",
    bill_depth_mm = "Bill depth (mm)",
    flipper_length_mm = "Flipper length (mm)",
    body_mass_g = "Body mass (g)",
  ) |> 
  mutate(sex = factor(sex, labels = c("Female", "Male"))) |> 
  tidyr::drop_na(sex) 
```

### 1. First setup with `set_blanket()`

The `set_blanket()` function should be run first. 

This sets the default style of plots with themes and colours etc. It can be customised. 

It should be run at the start of every script or quarto document. 

```{r}
set_blanket()
```

### 2. Each `gg_*` function wraps a geom

Each `gg_*` function wraps a `ggplot2::ggplot()` function with the associated `ggplot2::geom_*()` function. 

Almost every geom in ggplot2 is wrapped. 

Most aesthetics can be added directly as arguments. 

```{r}
penguins2 |>
  gg_point(
    x = flipper_length_mm,
    y = body_mass_g,
  )
```

### 3. A merged `col` argument to colour/fill by a variable

The colour and fill aesthetics of ggplot2 are merged into a single concept represented by the `col` argument. 

This combined aesthetic means that all colour outlines and all fill interiors should be coloured with the `col_palette` by the `col` variable. 

Use `colour = NA` or `fill = NA` to turn one of these off.

```{r, fig.asp=0.45}
penguins2 |>
  gg_boxplot(
    x = flipper_length_mm,
    y = island,
    col = sex, 
  )
```

### 4. A `facet` argument to facet by a variable

Users provide an unquoted `facet` variable to facet by. 

When `facet` is specified, the `facet_layout` will default to a `"wrap"` of the `facet` variable (if `facet2 = NULL`).

```{r}
penguins2 |>
  gg_histogram(
    x = flipper_length_mm,
    facet = species,
  )
```

### 5. A `facet2` argument to facet by a 2nd variable

Users can also provide an unquoted `facet2` variable to facet by.

When `facet2` is specified, the `facet_layout` will default to a `"grid"` of the `facet` variable (horizontally) by the `facet2` variable (vertically).

```{r, fig.asp=0.75}
penguins2 |>
  gg_histogram(
    x = flipper_length_mm,
    facet = species,
    facet2 = sex,
  )
```

### 6. Other aesthetics via `mapping` argument

Some aesthetics are not available via an argument (e.g. `alpha`, `size`, `shape`, `linetype` and `linewidth`).

These can be accessed via the `mapping` argument with aesthetics placed within `ggplot2::aes(...)`.

To customise associated scales/guides, `+` on the applicable ggplot2 layer. 

```{r}
penguins2 |> 
  gg_jitter(
    y = species, 
    x = flipper_length_mm, 
    col = species,
    mapping = aes(shape = species),
  ) +
  scale_shape_manual(values = rev(shape_pal()(3)))
```

### 7.  Prefixed arguments to customise x/y/col/facet

There are numerous arguments to customise plots that are prefixed by whether they relate to `x`, `y`, `col` or `facet`. 

For `x`, `y` and `col`, these relate to associated arguments within ggplot2 scales and guides. For `facet`, they relate to associated arguments within `ggplot2::facet_wrap` and `ggplot2::facet_grid`. 

Scales and guides associated with other other aesthetics can be customised by adding the applicable ggplot2 layer. 

```{r}
penguins2 |>
  gg_jitter(
    x = flipper_length_mm,
    y = body_mass_g,
    col = flipper_length_mm,
    x_breaks = scales::breaks_extended(n = 4, only.loose = TRUE),
    x_labels = \(x) stringr::str_sub(x, 1, 1),
    y_expand_limits = 1000,
    y_labels = label_number(big.mark = " "), 
    y_transform = "log10",
    col_label = "Flipper\nlength (mm)",
    col_steps = TRUE,
    col_breaks = \(x) quantile(x, seq(0, 1, 0.25)),
    col_palette = viridis::rocket(n = 9, direction = -1),
  )
```

### 8.  Smart `*_label` defaults for axis and legend titles

The `x_label`, `y_label` and `col_label` for the axis and legend titles can be manually specified with the applicable `*_label` argument (or `+ ggplot2::labs(...)`).

If not specified, they will first take any label attribute associated with the applicable variable. 

If none, they will then convert the variable name to a label name using the `label_to_case` function, which defaults to sentence case (i.e. `snakecase::to_sentence_case`). 

```{r, fig.asp=0.6}
penguins2 |>
  gg_freqpoly(
    x = flipper_length_mm,
    col = species,
  )
```

### 9. Other `ggplot2::geom_*` arguments via `...`

The `...` argument provides access to all other arguments in the `ggplot2::geom_*()` function. 

Common arguments to add include `colour`, `fill`, `alpha`, `linewidth`, `linetype`,  `size` and  `width`, which enables fixing of these to a particular value. 

Use the `ggplot2::geom_*` help to see what arguments are available. 

```{r}
penguins2 |>
  gg_smooth(
    x = flipper_length_mm,
    y = body_mass_g,
    col = sex, 
    col_palette = c("#003f5c", "#ffa600"),
    colour = "#bc5090", 
    linewidth = 1, 
    linetype = "dashed",
    alpha = 1, 
    se = TRUE, 
    level = 0.999, 
  ) 
```

### 10. A 'symmetric' continuous scale with easy opt-out

The `gg_*` function will generally create a 'symmetric' continuous y scale with `y_limits` the range of `y_breaks` and `y_expand` of `c(0, 0)`. The vice-versa will occur where a y orientation is guessed.

You can turn this 'symmetric' scale off using `*_limits = c(NA, NA)` (or `*_limits = c(0, NA)` for bars etc). For a date variable, use `lubridate::NA_Date_` instead of `NA`.

Note this 'symmetric' scale does not occur where the scale has a non-linear `*_transform` (or where the other positional scale is binned). 

```{r, echo=FALSE}
p1 <- penguins2 |>
  gg_pointrange(
    x = sex,
    y = flipper_length_mm,
    stat = "summary",
    position = position_dodge(),
    x_labels = \(x) str_sub(x, 1, 1),
    subtitle = "\ny_limits = NULL",
    mode = grey_mode_r()
  ) +
  labs(y = NULL)

p2 <- penguins2 |>
  gg_pointrange(
    x = sex,
    y = flipper_length_mm,
    stat = "summary",
    position = position_dodge(),
    x_labels = \(x) str_sub(x, 1, 1),
    y_limits = c(NA, NA),
    subtitle = "\ny_limits = c(NA, NA),",
    mode = grey_mode_r()
  ) +
  labs(y = NULL)

p3 <- penguins2 |>
  gg_col(
    x = sex,
    y = flipper_length_mm,
    stat = "summary",
    position = position_dodge(),
    width = 0.5,
    x_labels = \(x) str_sub(x, 1, 1),
    y_limits = c(0, NA),
    subtitle = "\ny_limits = c(0, NA),",
    mode = grey_mode_r()
  ) +
  labs(y = NULL)

p1 + p2 + p3
```

### 11. Ability to add multiple `geom_*` layers

Users can make plots with multiple layers by adding on `ggplot2::geom_*` layers. 

The `gg_*` function puts the aesthetic variables within the wrapped `ggplot` function. Therefore, these aesthetics will inherit to any subsequent layers added. 

It generally works best to add all aesthetics required to the `gg_*` function. 

As geoms are drawn in order, the `gg_*()` function should be appropriate to be the bottom geom of the plot. Subsequently, you may sometimes need to use `gg_blanket()` which defaults to a `blank` geom.

```{r, fig.asp = 0.5}
penguins2 |>
  group_by(species, sex) |>
  summarise(
    lower = quantile(body_mass_g, probs = 0.05),
    upper = quantile(body_mass_g, probs = 0.95),
    body_mass_g = mean(body_mass_g, na.rm = TRUE),
  ) |>
  labelled::copy_labels_from(penguins2) |> 
  gg_col(
    x = body_mass_g,
    xmin = lower, 
    xmax = upper,
    y = species,
    col = sex,
    position = position_dodge(),
    width = 0.75,
    x_expand_limits = 0,
  ) +
  geom_errorbar(
    width = 0.1, 
    position = position_dodge(width = 0.75),
    colour = lightness[1],
  )  
```

### 12.  Families of `*_mode_*` themes with legend variants 

`light_mode_*`, `grey_mode_*` and `dark_mode_*` theme families are provided with themes that differ based on legend placement with suffix `r` (right), `b` (bottom), `t` (top) and `n` (none). 

These are built for use with the `mode` argument in the `gg_*` functions, which has  side-effects for x and y orientations removing axis lines/ticks and gridlines. 

If you want to use a theme with no side-effects, then `+` on the theme instead.

```{r, fig.asp=0.65}
penguins2 |>
  gg_histogram(
    x = flipper_length_mm,
    col = species,
    title = "Penguin flipper length by species",
    subtitle = "Palmer Archipelago, Antarctica",
    caption = "Source: Gorman, 2020", 
    mode = grey_mode_t(),
  ) +
  theme(legend.title = element_blank())
```


### 13. Arguments to customise setup with `set_blanket()`

The `set_blanket` function sets customisable defaults for the:

* mode (i.e. a theme added with side-effects)
* geom colour (i.e. colour/fill where `col = NULL` for most geoms)
* annotation colour  (i.e. colour/fill where `col = NULL` for other geoms often used for annotation)
* discrete colour palette (and NA colour)
* continuous colour palette (and NA colour)
* theme (i.e. a theme added with no side-effects)

The `ggplot2::update_geom_defaults()` function can be used to further fine-tune geom defaults. 

`set_blanket()` also works on ggplot2 code. 

```{r}
set_blanket(
  mode = dark_mode_r(), 
  geom_colour = "#e7298a",
  annotate_colour = darkness[1],
  col_palette_d = c("#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e", 
                    "#e6ab02", "#a6761d", "#666666"), #RColorBrewer Dark2 
)

p1 <- penguins2 |>
  gg_point(
    x = flipper_length_mm, 
    y = body_mass_g,
    x_breaks = breaks_extended(n = 4, only.loose = TRUE),
  ) +
  geom_vline(xintercept = 200) +
  annotate("text", x = I(0.25), y = I(0.75), label = "Here")

p2 <- penguins2 |> 
  gg_histogram(
    x = flipper_length_mm,
    col = species,
    x_breaks = breaks_extended(n = 4, only.loose = TRUE),
  ) +
  geom_vline(xintercept = 200) +
  annotate("text", x = I(0.75), y = I(0.75), label = "Here")

p1 + p2

set_blanket()
```

### 14. A `gg_blanket()` function with `geom` flexibility

The package is driven by the `gg_blanket` function, which has a `geom` argument with `geom_blank` geom, stat and position defaults.

All other functions wrap this function with a fixed `geom`, and their own default `stat` and `position` arguments as per the applicable `geom_*` function.

You can print a `geom_*` function to identify the default `stat` and `position` arguments etc. 

```{r}
geom_histogram()

penguins2 |>
  gg_blanket(
    geom = "bar",
    stat = "bin",
    position = "stack",
    x = flipper_length_mm, 
    col = species,
  ) 
```
## Further information 

See the ggblanket [website](https://davidhodge931.github.io/ggblanket/index.html) for further information, including [articles](https://davidhodge931.github.io/ggblanket/articles/) and [function reference](https://davidhodge931.github.io/ggblanket/reference/index.html).
