---
title: "ggblanket"
author: "David Hodge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggblanket}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.asp = 0.6,
  out.width = "70%",
  dpi = 300
)
```

## Overview

ggblanket is a package of ggplot2 wrapper functions.

The primary objective is to **simplify ggplot2 visualisation**.

Secondary objectives relate to:

* Design: produce well-designed visualisation
* Alignment: align with ggplot2 and tidyverse
* Scope: cover much of what ggplot2 does.

Computational speed has been traded-off.

## How it works

1.  First setup with `set_blanket()`
2.  Each `gg_*` function wraps a geom
3.  A `col` aesthetic representing both colour and fill
4.  A `facet` argument to facet by a variable
5.  A `facet2` argument to facet by a 2nd variable
6.  Other aesthetics via `mapping` argument
7.  Prefixed arguments to customise x/y/col/facet
8.  Smart `*_title` defaults for axis and legend titles
9.  Other `ggplot2::geom_*` arguments via `...`
10.  `theme_lighter` and `theme_darker` themes 
11.  Side-effects to the `theme` based on the `theme_orientation`
12.  One `*_symmetric` continuous scale
13. Ability to add multiple `geom_*` layers
14. Arguments to customise setup with `set_blanket()`
15. A `gg_blanket()` function with `geom` flexibility

```{r setup}
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)
library(ggblanket)
library(palmerpenguins)
library(patchwork)
```

### 1. First setup with `set_blanket()`

The `set_blanket()` function should be run first. 

This sets the default style of plots with themes and colours etc. It can be customised. 

It should be run at the start of every script or quarto document. 

```{r}
set_blanket()
```

### 2. Each `gg_*` function wraps a geom

Each `gg_*` function wraps a `ggplot2::ggplot()` function with the associated `ggplot2::geom_*()` function. 

Almost every geom in ggplot2 is wrapped. 

Position related aesthetics can be added directly as arguments. 

```{r}
palmerpenguins::penguins |>
  gg_point(
    x = flipper_length_mm,
    y = body_mass_g,
  )
```

### 3.  A `col` aesthetic representing both colour and fill

In ggplot2, the `col` aesthetic is shorthand for the `colour` aesthetic.

In ggblanket, the `col` aesthetic represents *both* the `colour` *and* the `fill` aesthetic. 

The `col` argument is the variable to map the `colour` and `fill` aesthetics. 

```{r}
palmerpenguins::penguins |>
  tidyr::drop_na(sex) |> 
  gg_violin(
    x = flipper_length_mm,
    y = island,
    col = sex, 
  )
```

### 4. A `facet` argument to facet by a variable

Users provide an unquoted `facet` variable to facet by. 

When `facet` is specified, the `facet_layout` will default to a `"wrap"` of the `facet` variable (if `facet2 = NULL`).

```{r}
palmerpenguins::penguins |>
  gg_histogram(
    x = flipper_length_mm,
    facet = species,
  )
```

### 5. A `facet2` argument to facet by a 2nd variable

Users can also provide an unquoted `facet2` variable to facet by.

When `facet2` is specified, the `facet_layout` will default to a `"grid"` of the `facet` variable (horizontally) by the `facet2` variable (vertically).

```{r, fig.asp=0.75}
palmerpenguins::penguins |>
  gg_histogram(
    x = flipper_length_mm,
    facet = species,
    facet2 = sex,
  )
```

### 6. Other aesthetics via `mapping` argument

Some aesthetics are not available via an argument (e.g. `alpha`, `size`, `shape`, `linetype` and `linewidth`).

These can be accessed via the `mapping` argument using the `aes()` function. 

To customise the scales/guides of these other aesthetics, just `+` on the applicable ggplot2 layer. In some situations, you may need to override the colour used in the guide, or reverse the values in the relevant scale etc.

```{r}
palmerpenguins::penguins |> 
  tidyr::drop_na(sex) |> 
  gg_jitter(
    x = species, 
    y = flipper_length_mm, 
    col = island,
    mapping = aes(shape = sex),
  ) +
  guides_shape_grey()
```

### 7.  Prefixed arguments to customise x/y/col/facet

There are numerous arguments to customise plots that are prefixed by whether they relate to `x`, `y`, `col` or `facet`. 

For `x`, `y` and `col`, these relate to associated arguments within ggplot2 scales and guides. For `facet`, they relate to associated arguments within `ggplot2::facet_wrap` and `ggplot2::facet_grid`. 

Scales and guides associated with other other aesthetics can be customised by adding the applicable ggplot2 layer. 

```{r}
palmerpenguins::penguins |>
  gg_jitter(
    x = flipper_length_mm,
    y = body_mass_g,
    col = bill_length_mm,
    x_breaks_n = 4,
    x_title = "Flipper length",
    x_labels = \(x) paste0(x, " mm"),
    y_limits_include = 1000,
    y_labels = label_number(big.mark = " "), 
    y_transform = "sqrt",
    col_title = "Bill\nlength (mm)",
    col_breaks = \(x) quantile(x, seq(0, 1, 0.25)),
  )
```

### 8.  Smart `*_title` defaults for axis and legend titles

The `x_title`, `y_title` and `col_title` for the axis and legend titles can be manually specified with the applicable `*_title` argument (or `+ ggplot2::labs(...)`).

If not specified, they will first take any label attribute associated with the applicable variable. 

If none, they will then convert the variable name to a label name using the `titles_case` function, which defaults to sentence case (i.e. `snakecase::to_sentence_case`). 

```{r, fig.asp=0.6}
palmerpenguins::penguins |>
  gg_freqpoly(
    x = flipper_length_mm,
    col = species,
  )
```

### 9. Other `ggplot2::geom_*` arguments via `...`

The `...` argument provides access to all other arguments in the `ggplot2::geom_*()` function. 

Common arguments to add include `colour`, `fill`, `alpha`, `linewidth`, `linetype`,  `size` and  `width`, which enables fixing of these to a particular value. 

Use the `ggplot2::geom_*` help to see what arguments are available. 

```{r}
palmerpenguins::penguins |>
  gg_smooth(
    x = flipper_length_mm,
    y = body_mass_g,
    linewidth = 1, 
    linetype = "dashed",
    level = 0.999, 
    alpha = 1,
  ) 
```

### 10.  `theme_lighter` and `theme_darker` themes 

`theme_lighter` and `theme_darker`. These functions were built for use with the `theme` argument - and have flexibility to adjust colours, linewidths etc. 

```{r, fig.asp=0.65}
set_blanket(theme = theme_darker())
      
palmerpenguins::penguins |>
  gg_histogram(
    x = flipper_length_mm,
    col = species,
    title = "Penguin flipper length by species",
    subtitle = "Palmer Archipelago, Antarctica",
    caption = "Source: Gorman, 2020", 
  ) 

set_blanket()
```

### 11.  Side-effects to the `theme` based on the `perspective`

The `gg_*` function adds helpful side-effects to the theme based on the `perspective`, `axis_line_transparent`, `axis_ticks_transparent` and , `panel_grid_transparent` arguments. 

* Where `perspective = "x"`, the `gg_*` function will remove the y axis line/ticks and the x gridlines from the `theme` (by changing their colour to "transparent"). 
* Where `perspective = "y"`, the opposite will occur. 

If the `gg_*` guesses an incorrect `perspective`, then the user can change this. The user can change any of these side-effects by setting `axis_line_transparent`, `axis_ticks_transparent` or , `panel_grid_transparent` to `FALSE`.

Any `theme` used should be designed to anticipate the side-effects desired (i.e. include panel grid, axis line and axis ticks in both directions. 

This enables _one_ theme to be used across all plots whether the plot is vertical or horizontal. 

```{r}
p1 <- palmerpenguins::penguins |>
  gg_jitter(
    x = sex,
    y = bill_depth_mm,
  )

p2 <- palmerpenguins::penguins |>
  gg_jitter(
    x = bill_depth_mm,
    y = sex,
  ) 

p1 + p2
```

### 12.  One `*_symmetric` continuous scale

A `*_symmetric` continuous scale can be made where:

* `x_transform` and `y_transform` are NULL
* the `stat` is not "sf"
* if faceted, the relevant scale is "fixed" 

Both symmetric continuous scales cannot be made where the stat is not 'identity'.

By default, the gg_* function will make a `x_symmetric` scale if there is a y discrete axis and a x continuous axis. Otherwise, it will make a `y_symmetric` scale. 

A `y_symmetric` axis makes:

* the limits locked to the range of the `y_breaks`
* the `y_expand` default to `c(0, 0)`.

The vice versa occurs for an `x_symmetric` axis.

Use `*_symmetric = FALSE` to revert to the normal limits and expand defaults (i.e. the limits of the range of the data with `*_expand = c(0.05, 0.05)`).

Note all continuous scales ensure _all_ data is kept using `scales::oob_keep` - and, by default, are left unclipped (i.e. `coord_cartesian(clip = "off")`. 

### 13. Ability to add multiple `geom_*` layers

Users can make plots with multiple `ggplot2::geom_*` layers. 

The `gg_*()` geom layer will be the bottom geom layer of the plot, and each subsequent `geom_*()` layer is placed on top. 

Aesthetics added directly (e.g. `x`, `y` etc.) to the `gg_*()` function will inherit to later `geom_*()` layers, whereas those added to the `mapping` argument will not. 

```{r}
palmerpenguins::penguins |> 
  gg_violin(
    x = species, 
    y = bill_depth_mm,
    outliers = FALSE,
  ) +
  geom_jitter(shape = 19) 
```

The scales are built within the `gg_*()` function _without_ knowledge of later layers. The `gg_*()` function builds scales with regard to the `stat`, `position`, and aesthetics (that the geom understands) etc. So, in some situations, users will need to take care. 

```{r}
palmerpenguins::penguins |>
  tidyr::drop_na(sex) |> 
  group_by(species, sex) |>
  summarise(
    lower = quantile(bill_depth_mm, probs = 0.05),
    upper = quantile(bill_depth_mm, probs = 0.95),
    bill_depth_mm = mean(bill_depth_mm, na.rm = TRUE),
  ) |>
  labelled::copy_labels_from(palmerpenguins::penguins) |>
  gg_blanket(
    y = species,
    x = bill_depth_mm,
    xmin = lower, 
    xmax = upper,
    col = sex,
    col_border = TRUE, 
    x_limits_include = 0,
  ) +
  geom_col(
    width = 0.75,
    position = position_dodge(),
  ) +
  geom_errorbar(
    width = 0.1, 
    position = position_dodge(width = 0.75),
  ) 
```

### 14. Arguments to customise setup with `set_blanket()`

The `set_blanket` function sets customisable defaults for the:

* the geom defaults, including the colour/fill
* the colour/fill palettes for discrete, continuous and ordinal
* the theme, and how/what side-effects are to be applied
* the function to apply to a unspecified/unlabelled `x_title`, `y_title`, `col_title` etc.

Users should try to do as much as possible in these set-up functions, rather than continually adding arguments to their `gg_*` functions.

```{r}
set_blanket(
  theme = theme_lighter(
      base_size = 9,
      legend_position = "top",
      axis_line_linewidth = 0.25, 
      axis_line_colour = "#78909C",
      panel_grid_colour = "#C8D7DF",
      panel_background_fill = "#E8EFF2",
      panel_grid_linewidth = 0.25, 
  ),
  col = "tan", 
  col_palette_d = c("#003f5c", "#bc5090", "#ffa600", "#357BA2"),
)

p1 <- palmerpenguins::penguins |>
  gg_point(
    x = bill_depth_mm, 
    y = flipper_length_mm,
  ) 

p2 <- palmerpenguins::penguins |> 
  gg_bar(
    y = sex,
    col = species,
    position = "dodge",
  ) 

p1 + p2

set_blanket()
```

### 15. A `gg_blanket()` function with `geom` flexibility

The package is driven by the `gg_blanket` function, which has a `geom` argument with `ggplot2::geom_blank` defaults for `geom`, `stat` and `position`.

All other functions wrap this function with a fixed `geom`, and their own default `stat` and `position` arguments as per the applicable `geom_*` function.

This function can often be used with geoms that do not have an associated `gg_*` function.

```{r}
geom_spoke()

expand.grid(x = 1:10, y = 1:10) |>
  tibble() |>
  mutate(angle = runif(100, 0, 2*pi)) |>
  mutate(speed = runif(100, 0, sqrt(0.1 * x))) |>
  gg_blanket(
    geom = "spoke",
    x = x, 
    y = y,
    col = speed,
    mapping = aes(angle = angle, radius = speed),
  ) +
  geom_point()
```

## Further information 

See the ggblanket [website](https://davidhodge931.github.io/ggblanket/index.html) for further information, including [articles](https://davidhodge931.github.io/ggblanket/articles/) and [function reference](https://davidhodge931.github.io/ggblanket/reference/index.html).
