---
title: "ggblanket"
author: "David Hodge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggblanket}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.asp = 0.6,
  out.width = "70%",
  dpi = 300)
```

## Overview

ggblanket is a package of ggplot2 wrapper functions.

The primary objective is to **simplify ggplot2 visualisation**.

Secondary objectives relate to:

* Design: produce well-designed visualisation
* Alignment: align with ggplot2 and tidyverse
* Scope: cover much of what ggplot2 does.

Computational speed has been traded-off.

## How it works

1.  First setup with `set_blanket()`
2.  Each `gg_*` function wraps a geom
3.  A merged `col` argument to colour/fill by a variable
4.  A `facet` argument to facet by a variable
5.  A `facet2` argument to facet by a 2nd variable
6.  Smart `*_label` defaults for axis and legend titles
7.  Prefixed arguments to customise 
8.  Other `geom_*` arguments via `...`
9.  Families of `*_mode_*` themes with legend variants
10. Other aesthetics via `mapping`
11. A 'symmetric' continuous scale with easy opt-out
12. Ability to add multiple `geom_*` layers
13. A `gg_blanket()` function with `geom` flexibility
14. Arguments to customise setup with `set_blanket()`


```{r setup}
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)
library(ggblanket)
library(palmerpenguins)
library(patchwork)

penguins2 <- penguins |> 
  labelled::set_variable_labels(
    bill_length_mm = "Bill length (mm)",
    bill_depth_mm = "Bill depth (mm)",
    flipper_length_mm = "Flipper length (mm)",
    body_mass_g = "Body mass (g)",
  ) |> 
  mutate(sex = factor(sex, labels = c("Female", "Male"))) |> 
  tidyr::drop_na(sex) 
```

### 1. First setup with `set_blanket()`

The `set_blanket()` function should be run first. 

This sets the default style of plots with themes and colours etc. It can be customised. 

It should be run at the start of every script or quarto document. 

```{r}
set_blanket()
```

### 2. Each `gg_*` function wraps a geom

Each `gg_*` function wraps a `ggplot2::ggplot()` function with the associated `ggplot2::geom_*()` function. 

Almost every geom in ggplot2 is wrapped. 

Most aesthetics can be added directly as arguments. 

```{r}
penguins2 |>
  gg_point(
    x = flipper_length_mm,
    y = body_mass_g,
  )
```

### 3. A merged `col` argument to colour/fill by a variable

The colour and fill aesthetics of ggplot2 are merged into a single concept represented by the `col` argument. 

This combined aesthetic means that everything should be coloured according to it, i.e. all `colour` outlines and all `fill` interiors. 

Use `colour = NA` or `fill = NA` to turn one of these off.

```{r}
penguins2 |>
  gg_point(
    x = flipper_length_mm,
    y = body_mass_g,
    col = species, 
  )
```

```{r, fig.asp=0.45}
penguins2 |>
  gg_bar(
    position = "dodge",
    y = species, 
    col = sex, 
    width = 0.75,
  )
```

### 4. A `facet` argument to facet by a variable

Users provide an unquoted `facet` variable to facet by. 

When `facet` is specified (and `facet2 = NULL`), the `facet_layout` will default to a `"wrap"` of the `facet` variable.

```{r}
penguins2 |>
  gg_histogram(
    x = flipper_length_mm,
    facet = species,
  )
```

### 5. A `facet2` argument to facet by a 2nd variable

A `facet2` argument is also provided for extra functionality and flexibility. 

When `facet2` is specified, the `facet_layout` will default to a `"grid"` of the `facet` variable (horizontally) by the `facet2` variable (vertically).

```{r, fig.asp=0.75}
penguins2 |>
  gg_histogram(
    x = flipper_length_mm,
    facet = species,
    facet2 = sex,
  )
```

### 6.  Smart `*_label` defaults for axis and legend titles

The `x_label`, `y_label` and `col_label` for the axis and legend titles can be manually specified with the applicable `*_label` argument.

If not specified, they will first take any label attribute associated with the applicable variable. 

If none, they will then convert the variable name to a label name using the `label_to_case` function, which defaults to sentence case (i.e. `snakecase::to_sentence_case`). 

```{r, fig.asp=0.6}
penguins2 |>
  gg_freqpoly(
    x = flipper_length_mm,
    col = species,
  )
```

### 7.  Prefixed arguments to customise

ggblanket offers numerous arguments to customise that are prefixed by whether they relate to `x`, `y`, `col` or `facet`. 

For `x`, `y` and `col`, these mostly relate to associated arguments within ggplot2 scales and guides. For `facet`, they relate to associated arguments within `ggplot2::facet_wrap` and `ggplot2::facet_grid`. 

These prefixed arguments work nicely with auto-complete, if users ensure their settings support this.

```{r}
penguins2 |>
  gg_jitter(
    x = species,
    y = body_mass_g,
    col = flipper_length_mm,
    facet = sex,
    x_labels = \(x) stringr::str_sub(x, 1, 1),
    y_breaks = breaks_width(1000),
    y_expand_limits = 1000,
    y_labels = label_number(big.mark = " "), 
    y_transform = "log10",
    col_steps = TRUE,
    col_breaks = \(x) quantile(x, seq(0, 1, 0.25)),
    col_palette = viridis::magma(n = 9, direction = -1),
  )
```

### 8. Other `geom_*` arguments via `...`

The `...` argument provides access to all other arguments in the `geom_*()` function. 

Common arguments to add include `colour`, `fill`, `alpha`, `linewidth`, `linetype`,  `size` and  `width`.

Use the `geom_*` help to see what arguments are available. 

```{r}
penguins2 |>
  gg_smooth(
    x = flipper_length_mm,
    y = body_mass_g,
    col = sex, 
    col_palette = c("#003f5c", "#ffa600"),
    colour = "#bc5090", 
    linewidth = 1, 
    linetype = "dashed",
    alpha = 1, 
    se = TRUE, 
    level = 0.999, 
  ) 
```

### 9.  Families of `*_mode_*` themes with legend variants 

Three families of themes are provided: `light_mode_*`, `grey_mode_*` and `dark_mode_*`.

Each theme family provides 4 variants that differ based on legend placement with suffix `r` (right), `b` (bottom), `t` (top) and `n` (none). 

These are built for use with the `mode` argument in the `gg_*` functions, which has  side-effects for x and y orientations removing axis lines/ticks and gridlines). 

If you want to use a theme with no side-effects, then `+` the theme to the plot.

```{r, fig.asp=0.65}
penguins2 |>
  gg_histogram(
    x = flipper_length_mm,
    col = species,
    title = "Penguin flipper length by species",
    subtitle = "Palmer Archipelago, Antarctica",
    caption = "Source: Gorman, 2020", 
    mode = grey_mode_t(),
  ) +
  theme(legend.title = element_blank())
```

### 10. Other aesthetics via `mapping`

Some aesthetics are not available via an argument (e.g. `alpha`, `size`, `shape`, `linetype` and `linewidth`).

These can be accessed via the `mapping` argument with aesthetics placed within a `ggplot2::aes` function.

To customise associated scales/guides, `+` on the applicable ggplot2 layer. 

```{r}
penguins2 |> 
  gg_jitter(
    y = species, 
    x = flipper_length_mm, 
    col = species,
    mapping = aes(shape = species),
  ) +
  scale_shape_manual(values = rev(shape_pal()(3)))
```

### 11. A 'symmetric' continuous scale with easy opt-out

The `gg_*` function will generally create a 'symmetric' continuous y scale with (1.) `y_limits` the range of `y_breaks` and `y_expand` of `c(0, 0)`.

The vice-versa will occur where a y orientation is guessed.

You can turn this 'symmetric' scale off using `*_limits = c(NA, NA)` (or `*_limits = c(0, NA)` for bars etc). For a date variable, use `lubridate::NA_Date_` instead of `NA`.

Note this 'symmetric' scale does not occur where the scale has a non-linear `*_transform` (or where the other positional scale is binned). 

```{r, echo=FALSE}
p1 <- penguins2 |>
  gg_pointrange(
    x = sex,
    y = flipper_length_mm,
    stat = "summary",
    position = position_dodge(),
    x_labels = \(x) str_sub(x, 1, 1),
    subtitle = "\ny_limits = NULL",
    mode = light_mode_r(panel_grid_colour = "black")
  ) 

p2 <- penguins2 |>
  gg_pointrange(
    x = sex,
    y = flipper_length_mm,
    stat = "summary",
    position = position_dodge(),
    x_labels = \(x) str_sub(x, 1, 1),
    y_limits = c(NA, NA),
    subtitle = "\ny_limits = c(NA, NA),",
    mode = light_mode_r(panel_grid_colour = "black")
  ) 

p3 <- penguins2 |>
  gg_col(
    x = sex,
    y = flipper_length_mm,
    stat = "summary",
    position = position_dodge(),
    width = 0.5,
    x_labels = \(x) str_sub(x, 1, 1),
    y_limits = c(0, NA),
    subtitle = "\ny_limits = c(0, NA),",
    mode = light_mode_r(panel_grid_colour = "black")
  ) 

p1 + p2 + p3
```

### 12. Ability to add multiple `geom_*` layers

Users can make plots with multiple layers by `+`-ing on `ggplot2::geom_*` layers. 

The `gg_*` function puts the aesthetic variables within the wrapped `ggplot` function. Therefore, these aesthetics will inherit to any subsequent layers added. 

It generally works best to add all aesthetics required to the `gg_*` function. 

As geoms are drawn in order, the `gg_*()` function should be appropriate to be the bottom geom of the plot. 

Subsequently, you may sometimes need to use `gg_blanket()` which defaults to a `blank` geom.

```{r, fig.asp = 0.4}
penguins2 |>
  group_by(species) |>
  summarise(
    lower = quantile(body_mass_g, probs = 0.05),
    upper = quantile(body_mass_g, probs = 0.95),
    body_mass_g = mean(body_mass_g, na.rm = TRUE),
  ) |>
  labelled::copy_labels_from(penguins2) |> 
  gg_blanket(
    x = body_mass_g,
    xmin = lower, 
    xmax = upper,
    y = species,
    col = species,
    x_expand_limits = 0,
  ) +
  geom_col(
    colour = "#d3d3d3",
    fill = "#d3d3d3",
    width = 0.75,
  ) +
  geom_errorbar(
    width = 0.1, 
  ) 
```

### 13. A `gg_blanket()` function with `geom` flexibility

ggblanket is driven by the `gg_blanket` function, which has a `geom` argument with `geom_blank` defaults.

All other functions wrap this function with a fixed `geom`, and their own default `stat` and `position` arguments as per the applicable `geom_*` function.

You can print a `geom_*` function to identify the applicable `stat` and `position` arguments etc. 

Note this function can often be used with ggplot2 extension package geoms.

```{r}
geom_histogram()

penguins2 |>
  gg_blanket(
    geom = "bar",
    stat = "bin",
    position = "stack",
    x = flipper_length_mm, 
    col = species,
  ) 
```

### 14. Arguments to customise setup with `set_blanket()`

The `set_blanket` function sets customisable defaults for the:

* mode (i.e. a theme added with side-effects)
* geom colour (i.e. colour/fill where `col = NULL`)
* annotation colour  (i.e. colour/fill where `col = NULL` for geoms often used for annotation)
* discrete colour palette (and NA colour)
* continuous colour palette (and NA colour)
* theme (i.e. a theme added with no side-effects)

The `ggplot2::update_geom_defaults()` function can be used to further fine-tune geom defaults. 

`set_blanket()` also works on ggplot2 code. 

```{r}
set_blanket(
  mode = dark_mode_r(), 
  geom_colour = "#bc5090",
  annotate_colour = darkness[1],
  col_palette_d = c(purple, "tan", pink),
)

p1 <- penguins2 |>
  gg_point(
    x = flipper_length_mm, 
    y = body_mass_g,
    x_breaks = breaks_extended(3, only.loose = TRUE),
  ) +
  geom_vline(xintercept = 200) +
  annotate("text", x = I(0.25), y = I(0.75), label = "Here")

p2 <- penguins2 |> 
  gg_histogram(
    x = flipper_length_mm,
    col = species,
    x_breaks = breaks_extended(3, only.loose = TRUE),
  ) +
  geom_vline(xintercept = 200) +
  annotate("text", x = I(0.75), y = I(0.75), label = "Here")

p1 + p2
```

## Further information 

See the ggblanket [website](https://davidhodge931.github.io/ggblanket/index.html) for further information, including [articles](https://davidhodge931.github.io/ggblanket/articles/) and [function reference](https://davidhodge931.github.io/ggblanket/reference/index.html).
