---
title: "ggblanket"
author: "David Hodge"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ggblanket}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.asp = 0.6,
  out.width = "70%",
  dpi = 300
)
```

## Overview

ggblanket is a package of ggplot2 wrapper functions.

The primary objective is to **simplify ggplot2 visualisation**.

Secondary objectives relate to:

* Design: produce well-designed visualisation
* Alignment: align with ggplot2 and tidyverse
* Scope: cover much of what ggplot2 does.

Computational speed has been traded-off.

## How it works

1.  First setup with `set_blanket()`
2.  Each `gg_*` function wraps a geom
3.  A merged `col` argument to colour/fill by a variable
4.  A `facet` argument to facet by a variable
5.  A `facet2` argument to facet by a 2nd variable
6.  Other aesthetics via `mapping` argument
7.  Prefixed arguments to customise x/y/col/facet
8.  Smart `*_label` defaults for axis and legend titles
9.  Other `ggplot2::geom_*` arguments via `...`
10. Families of `*_mode_*` themes with legend variants
11. Helpful orientation defaults and side-effects 
12. Ability to add multiple `geom_*` layers
13. Arguments to customise setup with `set_blanket()`
14. A `gg_blanket()` function with `geom` flexibility

```{r setup}
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)
library(ggblanket)
library(palmerpenguins)
library(patchwork)

penguins2 <- penguins |> 
  labelled::set_variable_labels(
    bill_length_mm = "Bill length (mm)",
    bill_depth_mm = "Bill depth (mm)",
    flipper_length_mm = "Flipper length (mm)",
    body_mass_g = "Body mass (g)",
  ) |> 
  mutate(sex = factor(sex, labels = c("Female", "Male"))) |> 
  tidyr::drop_na(sex) 
```

### 1. First setup with `set_blanket()`

The `set_blanket()` function should be run first. 

This sets the default style of plots with themes and colours etc. It can be customised. 

It should be run at the start of every script or quarto document. 

```{r}
set_blanket()
```

### 2. Each `gg_*` function wraps a geom

Each `gg_*` function wraps a `ggplot2::ggplot()` function with the associated `ggplot2::geom_*()` function. 

Almost every geom in ggplot2 is wrapped. 

Position related aesthetics can be added directly as arguments. 

```{r}
penguins2 |>
  gg_point(
    x = flipper_length_mm,
    y = body_mass_g,
  )
```

### 3. A merged `col` argument to colour/fill by a variable

The colour and fill aesthetics of ggplot2 are merged into a single concept represented by the `col` argument. 

This combined aesthetic means that all colour outlines and all fill interiors should be coloured with the `col_palette` by the `col` variable. 

Use `colour = NA` or `fill = NA` to turn one of these off.

```{r, fig.asp=0.45}
penguins2 |>
  gg_boxplot(
    x = flipper_length_mm,
    y = island,
    col = sex, 
  )
```

### 4. A `facet` argument to facet by a variable

Users provide an unquoted `facet` variable to facet by. 

When `facet` is specified, the `facet_layout` will default to a `"wrap"` of the `facet` variable (if `facet2 = NULL`).

```{r}
penguins2 |>
  gg_histogram(
    x = flipper_length_mm,
    facet = species,
  )
```

### 5. A `facet2` argument to facet by a 2nd variable

Users can also provide an unquoted `facet2` variable to facet by.

When `facet2` is specified, the `facet_layout` will default to a `"grid"` of the `facet` variable (horizontally) by the `facet2` variable (vertically).

```{r, fig.asp=0.75}
penguins2 |>
  gg_histogram(
    x = flipper_length_mm,
    facet = species,
    facet2 = sex,
  )
```

### 6. Other aesthetics via `mapping` argument

Some aesthetics are not available via an argument (e.g. `alpha`, `size`, `shape`, `linetype` and `linewidth`).

These can be accessed via the `mapping` argument using the `aes()` function. 

To customise associated scales/guides, `+` on the applicable ggplot2 layer. In some situations, you may have to reverse the values in the relevant scale etc.

Note you can use `set_blanket(geom_colour = lightness[1], ...)` to make legend symbols black if needed. 

```{r}
penguins2 |> 
  gg_jitter(
    x = species, 
    y = flipper_length_mm, 
    col = island,
    mapping = aes(shape = island),
  ) 
```

### 7.  Prefixed arguments to customise x/y/col/facet

There are numerous arguments to customise plots that are prefixed by whether they relate to `x`, `y`, `col` or `facet`. 

For `x`, `y` and `col`, these relate to associated arguments within ggplot2 scales and guides. For `facet`, they relate to associated arguments within `ggplot2::facet_wrap` and `ggplot2::facet_grid`. 

Scales and guides associated with other other aesthetics can be customised by adding the applicable ggplot2 layer. 

```{r}
penguins2 |>
  gg_jitter(
    x = flipper_length_mm,
    y = body_mass_g,
    col = flipper_length_mm,
    x_breaks = scales::breaks_extended(n = 4, only.loose = TRUE),
    x_labels = \(x) stringr::str_sub(x, 1, 1),
    y_expand_limits = 1000,
    y_labels = label_number(big.mark = " "), 
    y_transform = "log10",
    col_label = "Flipper\nlength (mm)",
    col_steps = TRUE,
    col_breaks = \(x) quantile(x, seq(0, 1, 0.25)),
    col_palette = viridis::rocket(n = 9, direction = -1),
  )
```

### 8.  Smart `*_label` defaults for axis and legend titles

The `x_label`, `y_label` and `col_label` for the axis and legend titles can be manually specified with the applicable `*_label` argument (or `+ ggplot2::labs(...)`).

If not specified, they will first take any label attribute associated with the applicable variable. 

If none, they will then convert the variable name to a label name using the `label_to_case` function, which defaults to sentence case (i.e. `snakecase::to_sentence_case`). 

```{r, fig.asp=0.6}
penguins2 |>
  gg_freqpoly(
    x = flipper_length_mm,
    col = species,
  )
```

### 9. Other `ggplot2::geom_*` arguments via `...`

The `...` argument provides access to all other arguments in the `ggplot2::geom_*()` function. 

Common arguments to add include `colour`, `fill`, `alpha`, `linewidth`, `linetype`,  `size` and  `width`, which enables fixing of these to a particular value. 

Use the `ggplot2::geom_*` help to see what arguments are available. 

```{r}
penguins2 |>
  gg_smooth(
    x = flipper_length_mm,
    y = body_mass_g,
    col = sex, 
    col_palette = c("#003F5CFF", "#FFA600FF"),
    colour = "#BC5090FF", 
    linewidth = 1, 
    linetype = "dashed",
    alpha = 1, 
    se = TRUE, 
    level = 0.999, 
  ) 
```

### 10.  Families of `*_mode_*` themes with legend variants 

`light_mode_*`, `grey_mode_*` and `dark_mode_*` theme families are provided with variants that differ based on legend placement with suffix `r` (right), `b` (bottom), `t` (top) and `n` (none). 

These themes were built for use with the `mode` argument.

```{r, fig.asp=0.65}
penguins2 |>
  gg_histogram(
    x = flipper_length_mm,
    col = species,
    title = "Penguin flipper length by species",
    subtitle = "Palmer Archipelago, Antarctica",
    caption = "Source: Gorman, 2020", 
    mode = grey_mode_t() + theme(legend.title = element_blank()),
  ) 
```

### 11.  Helpful orientation defaults and side-effects 

If not specified, the `gg_*` function will guess the plot orientation (based on whether the y axis is discrete).

Where `x_orientation = TRUE`, it will:

* remove the y axis line/ticks and the x gridlines from the `mode` theme. 
* make the linear y continuous scale 'symmetric' if possible with (a) `y_limits` the range of `y_breaks` and (b) `y_expand = c(0, 0)`.

The vice versa will occur for when `y_orientation = TRUE`. 

Note you can avoid all this by `+`-ing your theme on to the `gg_*` output, and adding a `*_limits = c(NA, NA)` argument (or `*_limits = c(0, NA)` for bars). 

```{r}
p1 <- penguins2 |>
  gg_jitter(
    x = sex,
    y = bill_depth_mm,
    subtitle = "\nx_orientation = TRUE",
  )

p2 <- penguins2 |>
  gg_jitter(
    x = bill_depth_mm,
    y = sex,
    subtitle = "\ny_orientation = TRUE",
  ) 

p1 + p2
```

### 12. Ability to add multiple `geom_*` layers

Users can make plots with multiple layers by adding on `ggplot2::geom_*` layers. 

Aesthetics directly added in the `gg_*` function are globally inherited. Those added via the `mapping` argument are only available locally to the layer associated with the `gg_*` function.  

It works best to add as many of the globally inherited aesthetics as possible to the `gg_*` function. 

The `gg_*()` function should also be appropriate to be the bottom layer of the plot. You may sometimes need to use `gg_blanket()` which defaults to a `blank` geom.

```{r, fig.asp = 0.5}
penguins2 |>
  group_by(species, sex) |>
  summarise(
    lower = quantile(flipper_length_mm, probs = 0.05),
    upper = quantile(flipper_length_mm, probs = 0.95),
    flipper_length_mm = mean(flipper_length_mm, na.rm = TRUE),
  ) |>
  labelled::copy_labels_from(penguins2) |>
  gg_col(
    x = flipper_length_mm,
    xmin = lower, 
    xmax = upper,
    y = species,
    col = sex,
    position = position_dodge(),
    width = 0.75,
  ) +
  geom_errorbar(
    width = 0.1, 
    position = position_dodge(width = 0.75),
    colour = lightness[1],
  )  
```


### 13. Arguments to customise setup with `set_blanket()`

The `set_blanket` function sets customisable defaults for the:

* mode (i.e. a theme added with side-effects)
* geom colour (i.e. colour/fill where `col = NULL` for most geoms)
* annotation colour  (i.e. colour/fill where `col = NULL` for other geoms often used for annotation)
* discrete colour palette (and NA colour)
* continuous colour palette (and NA colour)
* theme (i.e. a theme added with no side-effects)

The `ggplot2::update_geom_defaults()` function can be used to further fine-tune geom defaults. 

`set_blanket()` also works on ggplot2 code. 

```{r}
set_blanket(
  mode = dark_mode_r(), 
  geom_colour = "#E7298AFF",
  annotate_colour = darkness[1],
  col_palette_d = c("#1B9E77FF", "#D95F02FF", "#7570b3FF", "#E7298AFF", "#66A61EFF", 
                    "#E6AB02FF", "#A6761DFF", "#666666FF"), #RColorBrewer Dark2 
)

p1 <- penguins2 |>
  gg_point(
    x = flipper_length_mm, 
    y = body_mass_g,
    x_breaks = breaks_extended(n = 4, only.loose = TRUE),
  ) +
  geom_vline(xintercept = 200) +
  annotate("text", x = I(0.25), y = I(0.75), label = "Here")

p2 <- penguins2 |> 
  gg_histogram(
    x = flipper_length_mm,
    col = species,
    x_breaks = breaks_extended(n = 4, only.loose = TRUE),
  ) +
  geom_vline(xintercept = 200) +
  annotate("text", x = I(0.75), y = I(0.75), label = "Here")

p1 + p2

set_blanket()
```

### 14. A `gg_blanket()` function with `geom` flexibility

The package is driven by the `gg_blanket` function, which has a `geom` argument with `ggplot2::geom_blank` defaults for `geom`, `stat` and `position`.

All other functions wrap this function with a fixed `geom`, and their own default `stat` and `position` arguments as per the applicable `geom_*` function.

This function can often be used with geoms that do not have an associated `gg_*` function.

```{r}
geom_spoke()

expand.grid(x = 1:10, y = 1:10) |>
  tibble() |>
  mutate(angle = runif(100, 0, 2*pi)) |>
  mutate(speed = runif(100, 0, sqrt(0.1 * x))) |>
  gg_blanket(
    geom = "spoke",
    x = x, 
    y = y,
    col = speed,
    mapping = aes(angle = angle, radius = speed),
  ) +
  geom_point()
```

## Further information 

See the ggblanket [website](https://davidhodge931.github.io/ggblanket/index.html) for further information, including [articles](https://davidhodge931.github.io/ggblanket/articles/) and [function reference](https://davidhodge931.github.io/ggblanket/reference/index.html).
