---
title: "Go further"
author: "David Hodge"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.asp = 0.6,
  out.width = "70%",
  dpi = 300
  )
```

## Overview

This article will demonstrate a random assortment of content, including some of which is more advanced.  

1. Change the `stat` of the layer
2. Change the `position` of the layer
3. Use `+ *_mode_*` where axis-lines and gridlines are not as wanted
4. Reorder and/or reverse categorical variables
5. Drop unused categorical variable values
6. Transform scales to `"log"` etc
7. Correct the default orientation
8. Avoid the 'symmetric' scale
9. Avoid the mode side-effects
10. Change the `*_position` of positional axes 
11. Add contrasting dark or light text on polygons
12. Use opacity with to emphasise/de-emphasise 
13. Do the equivalent of setting limits
14. Use delayed evaluation
15. Rescale a diverging col scale
16. Add a legend within the panel

```{r setup}
library(dplyr)
library(tidyr)
library(forcats)
library(stringr)
library(ggplot2)
library(scales)
library(ggblanket)
library(patchwork)
library(palmerpenguins)

set_blanket()

penguins2 <- penguins |> 
  labelled::set_variable_labels(
    bill_length_mm = "Bill length (mm)",
    bill_depth_mm = "Bill depth (mm)",
    flipper_length_mm = "Flipper length (mm)",
    body_mass_g = "Body mass (g)",
  ) |> 
  mutate(sex = factor(sex, labels = c("Female", "Male")))
```

### 1. Change the `stat` of the layer

The default `stat` of each `gg_*` function can be changed. 

```{r}
penguins2 |>
  gg_pointrange(
    stat = "summary", 
    x = species,
    y = flipper_length_mm, 
  )
```

```{r}
library(ggforce)

ggplot2::economics |>
  slice_head(n = 35) |> 
  gg_path(
    stat = "bspline", n = 100,
    x = date, 
    y = unemploy,
    y_label = "Unemployment",
    linewidth = 1,
  ) 
```

### 2. Change the `position` of the layer

The default `position` of each `gg_*` function can be changed. 

```{r}
penguins2 |> 
  gg_point(
    position = ggbeeswarm::position_quasirandom(),
    x = sex, 
    y = flipper_length_mm,
    col = sex,
  ) +
  theme(legend.position = "none")
```

### 3. Use `+ *_mode_*` where axis-lines and gridlines are not as wanted

Sometimes the plot might guess the removal of gridlines and axis-line/ticks incorrectly. In these situations, you can `+ *_mode_*` on to the plot, and then remove whatever you want. 

```{r}
msleep |> 
  gg_point(
    x = bodywt, 
    y = brainwt,
    col = vore,
    col_labels = str_to_sentence,
    x_transform = "log10",
    y_transform = "log10",
  ) +
  light_mode_r() +
  guides(
    x = guide_axis_logticks(),
    y = guide_axis_logticks(),
  ) 
```

### 4. Reorder and/or reverse categorical variables

ggblanket requires unquoted variables only for `x`, `y`, `col`, `facet`, `facet2` and  `alpha`. You can often manipulate the data prior to plotting to achieve what you want (e.g. using `tidyr::drop_na`, `forcats::fct_rev` and/or `forcats::fct_reorder`).

```{r}
p1 <- diamonds |>
  count(color) |>
  gg_col(
    x = n,
    y = color,
    width = 0.75,
    x_labels = \(x) x / 1000,
    x_label = "Count (thousands)", 
    subtitle = "\nDefault order"
  )

p2 <- diamonds |>
  count(color) |>
  mutate(color = fct_rev(fct_reorder(color, n))) |>
  gg_col(
    x = n,
    y = color,
    width = 0.75,
    x_labels = \(x) x / 1000,
    x_label = "Count (thousands)", 
    subtitle = "\nRe-orderered"
  ) 

p1 + p2
```

### 5. Drop unused categorical variable values

ggblanket keeps unused factor levels in the plot. If users wish to drop unused levels they should likewise do it in the data prior to plotting using `forcats::fct_drop`. 

```{r}
p1 <- diamonds |> 
  count(color) |>
  filter(color %in% c("E", "G", "I")) |>
  gg_col(
    x = n,
    y = color,
    width = 0.75,
    x_labels = \(x) x / 1000,
    x_label = "Count (thousands)", 
    subtitle = "\nUnused levels kept",
  )

p2 <- diamonds |> 
  count(color) |>
  filter(color %in% c("E", "G", "I")) |>
  mutate(color = forcats::fct_drop(color)) |> 
  gg_col(
    x = n,
    y = color,
    width = 0.75,
    x_labels = \(x) x / 1000,
    x_label = "Count (thousands)", 
    subtitle = "\nUnused levels dropped",
  )

p1 + p2
```

### 6. Transform scales to `"log"` etc

Transform objects (e.g. `transform_log()` or character strings of these can be used to transform scales - including combining these.

```{r, fig.asp=0.75}
p1 <- pressure |>
  gg_point(
    x = temperature, 
    y = pressure, 
    x_breaks = breaks_extended(n = 4),
    y_breaks = breaks_pretty(n = 4),
    subtitle = "\nDefault",
  )

p2 <- pressure |>
  gg_point(
    x = temperature, 
    y = pressure, 
    x_breaks = breaks_extended(n = 4),
    y_breaks = breaks_pretty(n = 4),
    y_transform = "reverse", 
    subtitle = "\nReverse",
  )

p3 <- pressure |>
  gg_point(
    x = temperature, 
    y = pressure,
    x_breaks = breaks_extended(n = 4),
    y_transform = "log10",
    subtitle = "\nLog10", 
  )

p4 <- pressure |>
  gg_point(
    x = temperature, 
    y = pressure, 
    x_breaks = breaks_extended(n = 4),
    y_transform = c("log10", "reverse"),
    subtitle = "\nLog10 & Reverse",
  )

(p1 + p2) / (p3 + p4)
```

### 7. Correct the default orientation

The `gg_*` function guesses the `*_orientation` of the plot to determine how to make continuous axes and what side-effects to have on the provided mode. If it guesses incorrectly, use either the `x_orientation` or `y_orientation` argument.

```{r}
p1 <- penguins2 |>
  gg_point(
    x = bill_depth_mm,
    y = bill_length_mm,
    subtitle = "\nDefault orientation",
  )

p2 <- penguins2 |>
  gg_point(
    x = bill_depth_mm,
    y = bill_length_mm,
    y_orientation = TRUE,
    subtitle = "\nCorrected orientation",
  ) 

p1 + p2
```

### 8. Avoid the 'symmetric' scale

Symmetric scales can be turned off or on using `*_symmetric` arguments. 

```{r}
p1 <- penguins2 |>
  gg_pointrange(
    x = sex,
    y = bill_length_mm,
    stat = "summary",
    position = position_dodge(),
    x_labels = \(x) str_sub(x, 1, 1),
    subtitle = "\ny_symmetric = NULL",
  ) +
  labs(y = NULL)

p2 <- penguins2 |>
  gg_pointrange(
    x = sex,
    y = bill_length_mm,
    stat = "summary",
    position = position_dodge(),
    x_labels = \(x) str_sub(x, 1, 1),
    y_symmetric = FALSE,
    subtitle = "\ny_symmetric = FALSE,",
  ) +
  labs(y = NULL)

p3 <- penguins2 |>
  gg_col(
    x = sex,
    y = bill_length_mm,
    stat = "summary",
    position = position_dodge(),
    width = 0.5,
    x_labels = \(x) str_sub(x, 1, 1),
    y_symmetric = FALSE,
    subtitle = "\ny_symmetric = FALSE,",
  ) +
  labs(y = NULL)

p1 + p2 + p3
```

### 9. Avoid the mode side-effects

Where `*_orientation = TRUE`, it will remove the relevant axis line/ticks and the gridlines from the mode theme. 

But you can avoid this by `+`-ing your theme on to the plot, instead of adding it to the `mode` argument.

```{r}
p1 <- penguins2 |>
  gg_jitter(
    x = sex,
    y = bill_depth_mm,
    subtitle = "\nmode = light_more_r()",
  )

p2 <- penguins2 |>
  gg_jitter(
    x = sex,
    y = bill_depth_mm,
    subtitle = "\n+ light_more_r()",
  ) +
  light_mode_r()

p1 + p2
```

### 10. Change the `*_position` of positional axes 

Positional axes can be changed using `*_position`.

Note that for `x_position = "top"`, a caption must be added or modified to make this work nicely with a `*_mode_*` theme. 

```{r}
economics |>
  gg_line(
    x = date,
    y = unemploy,
    col = date,
    y_position = "right",
    x_position = "top",
    caption = "", 
    title = "Unemployment",
    subtitle = "1967\u20132015",
  ) 
```

### 11. Add contrasting dark or light text on polygons

Text on polygons can be coloured with a dark or light colour that is determined based on the underlying polygon colour using `!!!aes_contrast()`. This method was developed by Teun van den Brand (@teunbrand).  

```{r}
penguins2 |>
  count(species, sex) |>
  gg_col(
    x = sex,
    y = n, 
    col = species,
    position = position_dodge2(preserve = "single"),
    width = 0.75,
  ) +
  geom_text(
    mapping = aes(label = n, !!!aes_contrast()),
    position = position_dodge2(width = 0.75, preserve = "single"),
    vjust = 1.33,
    show.legend = FALSE,
  )
```

```{r}
penguins2 |>
  count(species, sex) |>
  gg_col(
    position = position_dodge2(preserve = "single"),
    x = n,
    y = sex,
    col = species,
    width = 0.75,
    mode = dark_mode_r(),
  ) +
  geom_text(
    mapping = aes(label = n, !!!aes_contrast("dark")),
    position = position_dodge2(width = 0.75, preserve = "single"),
    hjust = 1.25,
    show.legend = FALSE,
  )
```

### 12. Use opacity to emphasise/de-emphasise 

Use opacity with `mapping = aes(alpha = ...)`, `col = ...`, and `col_palette` to emphasise/de-emphasise parts of the visualisation. 

```{r, fig.asp=0.75}
p1 <- penguins2 |> 
  drop_na(sex) |> 
  gg_density(
    x = flipper_length_mm,
    x_breaks = breaks_extended(n = 4),
    y_breaks = breaks_pretty(n = 4),
    colour = alpha(orange, 0.33),
    fill = orange,
    alpha = 0.1,
    subtitle = "\nOpacity fill 0.1 & outline 0.33"
  )

p2 <- penguins2 |> 
  drop_na(sex) |> 
  gg_density(
    x = flipper_length_mm,
    col = sex,
    x_breaks = breaks_extended(n = 4),
    y_breaks = breaks_pretty(n = 4),
    col_palette = c(orange, "#78909C"),
    subtitle = "\ncol variable"
  )

p3 <- penguins2 |> 
  drop_na(sex) |> 
  gg_density(
    x = flipper_length_mm,
    col = sex, 
    mapping = aes(alpha = sex),
    x_breaks = breaks_extended(n = 4),
    y_breaks = breaks_pretty(n = 4),
    col_palette = c(orange, "#78909C"),
    subtitle = "\ncol and alpha variable",
  ) +
  guides(colour = "none", fill = "none", alpha = "none") + 
  scale_alpha_manual(values = c(0.5, 0.1))

p4 <- penguins2 |> 
  drop_na(sex) |> 
  gg_density(
    x = flipper_length_mm,
    col = sex, 
    mapping = aes(alpha = sex),
    x_breaks = breaks_extended(n = 4),
    y_breaks = breaks_pretty(n = 4),
    col_palette = c(orange, alpha("#78909C", 0.25)),
    subtitle = "\ncol and alpha variable & outline opacity",
  ) +
  scale_alpha_manual(values = c(0.5, 0.1))

(p1 + p2) / (p3 + p4)
```

### 13. Do the equivalent of setting limits

There are no `*_limits` arguments in ggblanket. 

Instead, users should use a combination of filtering the data, adding `*_expand_limits` and `coord = coord_cartesian(xlim = ..., ylim = ...)` arguments etc. 

```{r}
#To Zoom out, se *_expand_limits:
penguins |>
  gg_smooth(
    x =  body_mass_g, 
    y = bill_depth_mm,
    se = TRUE,
    x_expand_limits = c(0),
    y_expand_limits = c(10, 25),
  ) 
```
```{r}
#To zoom-in when the stat equals "identity", use dplyr::filter
penguins |>
  filter(between(bill_depth_mm, 0, 15)) |>
  gg_point(
    x =  bill_depth_mm, 
    y = body_mass_g,
  ) 
```



```{r}
#To zoom-in when the stat does _not_ equal "identity", use coord_cartesian 
#Then either recreate the breaks, or turn off the symmetric axis 
penguins |>
  gg_smooth(
    x =  body_mass_g, 
    y = bill_depth_mm,
    se = TRUE,
    coord = coord_cartesian(ylim = c(14.8, 15)), 
    # y_symmetric = FALSE,
    # y_breaks = scales::breaks_width(0.05), #recreate breaks
  ) 
```

### 14. Use delayed evaluation

 The `mapping` argument can be used for delayed evaluation with the `ggplot2::after_stat` function.

```{r, fig.asp=0.55}
penguins2 |>
  gg_histogram(
    x = flipper_length_mm,
    mapping = aes(y = after_stat(density)),
    facet = species,
  )
```

```{r, fig.asp=0.55}
faithfuld |>
  gg_contour(
    x = waiting,
    y = eruptions,
    z = density,
    mapping = aes(colour = after_stat(level)),
    bins = 8,
  )
```

### 15. Rescale a diverging col scale

Use `col_rescale` to rescale a diverging scale around a central point. 

```{r}
rescale_vctr <- sort(c(range(mpg$cty), 15))

mpg  |>  
  gg_point(
    x = displ, 
    y = hwy, 
    col = cty,
    col_palette = c(navy, teal, "white", orange, red),
    col_rescale = scales::rescale(rescale_vctr),
    col_breaks = scales::breaks_width(5),
    mode = dark_mode_r(),
  )
```

### 16. Add a legend within the panel

```{r}
set_blanket()

penguins2 |> 
  gg_histogram(
    x = flipper_length_mm,
    col = species,
    mode = light_mode_r(),
  ) +
  theme(legend.position = "inside") +
  theme(legend.position.inside = c(1, 0.975)) +
  theme(legend.justification = c(1, 1)) 
```


