---
title: "Extensions"
author: "David Hodge"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.asp = 0.6,
  out.width = "70%",
  dpi = 300
  )
```

## Overview

This article will demonstrate some fantastic ggplot2 extension packages. 

Note, an alternative is to use ggplot2 code directly instead. This may be easier depending on your familiarity with ggblanket/ggplot2 and the relevant extension package. The `set_blanket()` function used to get a lot of the ggblanket style also works with ggplot2.

```{r setup}
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)
library(ggblanket)
library(patchwork)
library(palmerpenguins)

set_blanket()

penguins2 <- penguins |> 
  labelled::set_variable_labels(
    bill_length_mm = "Bill length (mm)",
    bill_depth_mm = "Bill depth (mm)",
    flipper_length_mm = "Flipper length (mm)",
    body_mass_g = "Body mass (g)",
  ) |> 
  mutate(sex = factor(sex, labels = c("Female", "Male")))
```

### patchwork

The patchwork package enables plots to be patched together. 

A convenient hack to add a panel centred title can be to add a `facet_wrap` layer with a character title string for the `facets` argument. 

Otherwise you will need to modify margins

```{r, fig.asp=0.9}
p1 <- mtcars |> 
  gg_point(
    x = mpg, 
    y = disp,
    y_breaks = breaks_pretty(),
  ) +
  facet_wrap(~"Plot 1")

p2 <- mtcars |> 
  gg_boxplot(
    x = gear, 
    y = disp, 
    group = gear, 
    width = 0.5,
    y_breaks = breaks_pretty(),
  ) +
  facet_wrap(~"Plot 2")

p3 <- mtcars |> 
  gg_smooth(
    x = disp, 
    y = qsec, 
    x_breaks = breaks_pretty(n = 4), 
    y_breaks = breaks_pretty(),
  ) +
  facet_wrap(~"Plot 3")

(p1 + p2) / p3 +
  plot_annotation(
    title = "A collection of plots patched together",
    subtitle = "Works so nice!",
    theme = light_mode_r() +
      theme(
        plot.title = element_text(margin = margin(t = -11)),
        plot.subtitle = element_text(margin = margin(t = 5.5)),
      )
  )
```

### ggtext

The ggtext package support using markdown in text. 

```{r, fig.asp=0.75}
penguins2 |>
  gg_point(
    x = bill_depth_mm,
    y = bill_length_mm,
    col = species,
    title = "***Pygoscelis*** *penguin* bill **lengths** and depths",
    subtitle = "<span style = 'color: #0095A8 ;'>**Adelie**</span>, 
       <span style = 'color:#FF7043;'>**Chinstrap**</span>, 
       *and* <span style = 'color:#112E51;'>**Gentoo**</span>", 
    mode = light_mode_n()
  ) +
  theme(plot.title = ggtext::element_markdown()) +
  theme(plot.subtitle = ggtext::element_markdown())
```

### showtext

The showtext package enables the use of different fonts in plots. The `*_mode_*` theme functions provide a `family` argument.

```{r, fig.asp = 0.7, fig.showtext = TRUE}
head(sysfonts::font_families_google())

sysfonts::font_add_google("Covered By Your Grace", "grace")
sysfonts::font_add_google('Roboto Slab', 'roboto_slab')
sysfonts::font_add_google('Syne Mono', 'syne')

showtext::showtext_auto(enable = TRUE)

penguins2 |>
  gg_point(
    x = flipper_length_mm,
    y = body_mass_g,
    col = sex,
    facet = species,
    title = "penguins2 body mass by flipper length",
    subtitle = "Palmer Archipelago, Antarctica",
    caption = "Source: Gorman, 2020",
    mode = light_mode_r(family = "grace")
  ) +
  theme(
    plot.title = element_text(family = "roboto_slab"),
    plot.subtitle = element_text(family = "syne")
  )

showtext::showtext_auto(enable = FALSE)
```

### gghighlight

The gghighlight package enables geoms or parts thereof to be highlighted. 

```{r}
penguins2 |>
  gg_point(
    x = flipper_length_mm,
    y = body_mass_g,
  ) +
  gghighlight::gghighlight(body_mass_g >= 5000)
```

The `gg_*` function builds the scale for the data that it thinks will be within the panel. Therefore in some situations, this must be taken into account to ensure the scale builds correctly. 

```{r}
penguins2 |>
  gg_histogram(
    x = flipper_length_mm,
    col = species,
    x_breaks = breaks_pretty(4),
  ) + #build the scale for all data
  facet_wrap(~species) + #then add facet layer
  gghighlight::gghighlight() 
```

### ggforce

The ggforce package includes numerous extra geoms, stats etc.

```{r}
library(ggforce)
geom_bspline()

economics |>
  slice_head(n = 35) |> 
  gg_path(
    stat = "bspline", n = 100, type = "clamped",
    x = date, 
    y = unemploy,
    linewidth = 1,
    y_label = "Unemployment",
  ) 
```

```{r}
geom_mark_hull()

penguins2 |>
  gg_blanket(
    geom = "mark_hull",
    x = flipper_length_mm,
    y = body_mass_g,
    col = species,
    coord = coord_cartesian(),
  ) +
  geom_point()
```

### ggrepel

The ggrepel package can be used to neatly avoid overlapping labels. 

```{r}
library(ggrepel)
geom_text_repel()

mtcars |>
  tibble::rownames_to_column("car") |>
  filter(wt > 2.75, wt < 3.45) |>
  gg_point(
    x = wt, 
    y = mpg, 
    label = car, 
  ) +
  geom_text_repel() 
```

### ggbeeswarm

The ggbeeswarm package enables beeswarm plots.

```{r}
library(ggbeeswarm)
geom_beeswarm()
geom_quasirandom()

p1 <- penguins2 |> 
  gg_point(
    position = position_beeswarm(),
    x = sex, 
    y = flipper_length_mm,
    col = sex,
    x_labels = \(x) str_to_upper(str_sub(x, 1, 1)),
    subtitle = "\nBeeswarm",
    mode = light_mode_n(),
  ) 

p2 <- penguins2 |> 
  gg_point(
    position = ggbeeswarm::position_quasirandom(),
    x = sex, 
    y = flipper_length_mm,
    col = sex,
    x_labels = \(x) str_to_upper(str_sub(x, 1, 1)),
    subtitle = "\nQuasirandom",
    mode = light_mode_n(),
  ) 

p1 + p2
```

### ggridges

The ggridges package enables ridgeline plots. 

```{r}
library(ggridges)
geom_density_ridges()

ggridges::Catalan_elections |>
  rename_with(snakecase::to_snake_case) |>
  mutate(year = factor(year)) |>
  gg_blanket(
    geom = "density_ridges",
    stat = "density_ridges",
    coord = coord_cartesian(),
    x = percent,
    y = year,
    col = option,
    colour = "white",
    x_limits = c(0, 100),
    y_expand = expansion(c(0, 0.125)),
    alpha = 0.9,
  )
```

### ggdist

The ggdist package enables the visualisation of uncertainty. 

```{r}
library(ggdist)
library(distributional)
geom_slabinterval()

set.seed(123)

data.frame(
  distribution = c("Gamma(2,1)", "Normal(5,1)", "Mixture"),
  values = c(
    dist_gamma(2, 1),
    dist_normal(5, 1),
    dist_mixture(
      dist_gamma(2, 1),
      dist_normal(5, 1),
      weights = c(0.4, 0.6))
  )) |> 
  gg_blanket(
    geom = "slabinterval",
    stat = "slabinterval",
    y = distribution,
    mapping = aes(dist = values),
    colour = blue,
    fill = blue,
    alpha = 0.6,
    y_expand = c(0.05, 0.05),
    mode = light_mode_n(),
    # show_slab = FALSE,
    # show_interval = FALSE,
    # side = "both",
  )
```

```{r}
geom_lineribbon()

set.seed(123)

tibble(x = 1:10) |> 
  group_by(across(everything()))  |> 
  do(tibble(y = rnorm(100, .$x))) |> 
  median_qi(.width = c(0.5, 0.8, 0.95)) |>
  mutate(.width = factor(.width)) |>
  gg_blanket(
    geom = "lineribbon",
    x = x,
    y = y,
    ymin = .lower,
    ymax = .upper,
    col = .width,
    colour = viridis::mako(n = 9)[c(1)],
    col_legend_rev = TRUE,
    col_palette = viridis::mako(n = 9)[c(4,6,9)],
  )
```

### ggdensity

The ggdensity package provides visualizations of density estimates.  

```{r}
library(ggdensity)
geom_hdr()

set.seed(123)

data.frame(
  x = rnorm(1000),
  y = rnorm(1000),
  z = c(rep("A", times = 500), rep("B", times = 500))
) |>
  gg_blanket(
    geom = "hdr", 
    stat = "hdr",
    x = x, 
    y = y, 
    colour = NA,
    y_limits = c(NA, NA),
  ) +
  guides(colour = FALSE, fill = FALSE)
```

```{r}
set.seed(123)

data.frame(
  x = rnorm(1000),
  y = rnorm(1000),
  z = c(rep("A", times = 500), rep("B", times = 500))
) |>
  gg_blanket(
    geom = "hdr", 
    stat = "hdr",
    x = x, 
    y = y, 
    col = z,
    facet = z,
    colour = NA,
    y_limits = c(NA, NA),
  ) 
```

### ggpattern

```{r}
library(ggpattern)

penguins2 |>
  group_by(species, sex) |>
  summarise(across(body_mass_g, \(x) mean(x))) |>
  tidyr::drop_na() |> 
  labelled::copy_labels_from(penguins2) |> 
  gg_blanket(
    geom = "col_pattern",
    position = "dodge",
    y = species,
    x = body_mass_g,
    col = sex,
    mapping = aes(pattern = sex),
    width = 0.75,
  ) 
```


### paletteer

The paletteer package provides access to most of the palettes within R. 

```{r}
library(paletteer)

p1 <- penguins2 |>
  gg_point(
    x = flipper_length_mm,
    y = body_mass_g,
    col = species,
    col_palette = paletteer_d("RColorBrewer::Dark2"),
    x_breaks = breaks_pretty(3),
    mode = light_mode_n(),
  ) 

p2 <- penguins2 |>
  gg_point(
    x = flipper_length_mm,
    y = body_mass_g,
    col = bill_depth_mm,
    col_palette = paletteer_c("viridis::rocket", 9),
    x_breaks = breaks_pretty(3),
    mode = light_mode_n(),
  ) 

p1 + p2
```


### ggblend
 
The ggblend package provides blending of colours. You must use a graphics device that supports blending.

```{r, fig.asp=0.5}
p1 <- penguins2 |>
  gg_density(
    x = flipper_length_mm,
    col = species,
    y_limits = c(0, NA),
    mode = light_mode_n(),
  ) +
  facet_wrap(~"Normal") +
  ggeasy::easy_remove_x_axis() +
  ggeasy::easy_remove_y_axis()

p2 <- penguins2 |>
  gg_blanket(
    stat = "density",
    x = flipper_length_mm,
    col = species,
    y_limits = c(0, NA),
    mode = light_mode_n(),
  ) +
  facet_wrap(~"Blended") +
  geom_density(alpha = 0.6) |> ggblend::blend(blend = "multiply") +
  ggeasy::easy_remove_x_axis() +
  ggeasy::easy_remove_y_axis()

p1 + p2
```

### ggh4x

The ggh4x package includes enhanced facet functionality.

```{r}
iris |> 
  rename_with(\(x) snakecase::to_snake_case(x)) |> 
  mutate(across(species, \(x) str_to_sentence(x))) |> 
  mutate(size = if_else(species == "Setosa", "Short Leaves", "Long Leaves")) |> 
  gg_point(
    x = sepal_width, 
    y = sepal_length, 
    x_breaks = breaks_pretty(n = 3),
  ) +
  ggh4x::facet_nested(
    cols = vars(size, species), 
    nest_line = TRUE, 
    solo_line = TRUE,
  ) +
  theme(strip.text.x = element_text(margin = margin(t = 2.5, b = 5)))
```

### ggeasy

The ggeasy package provides a lot of support for easily modifying themes. Removing axes and gridlines is often useful. 

```{r}
library(ggeasy)

penguins2 |>
  gg_jitter(
    x = species, 
    y = body_mass_g, 
    col = sex, 
  ) +
  light_mode_t() +
  easy_remove_y_gridlines() +
  easy_remove_x_axis(what = c("line", "ticks")) +
  easy_remove_legend_title()
```


