---
title: "Extensions"
author: "David Hodge"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.asp = 0.6,
  out.width = "70%",
  dpi = 300
  )
```

## Overview

This article will demonstrate some fantastic ggplot2 extension packages. 

```{r setups}
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(stringr)
library(ggblanket)
library(palmerpenguins)

set_blanket()
```

### patchwork

The patchwork package enables plots to be patched together. 

A convenient hack to add a panel centred title can be to add a `facet_wrap` layer with a character title string for the `facets` argument. 

Otherwise you will need to modify margins

```{r, fig.asp=0.9}
library(patchwork)

p1 <- mtcars |> 
  gg_point(
    x = mpg, 
    y = disp,
    y_labels = replace_seq,
  ) +
  facet_wrap(~"Plot 1")

p2 <- mtcars |> 
  gg_boxplot(
    x = gear, 
    y = disp, 
    group = gear, 
    width = 0.5,
    y_labels = replace_seq,
  ) +
  facet_wrap(~"Plot 2")

p3 <- mtcars |> 
  gg_smooth(
    x = disp, 
    y = qsec, 
    x_breaks = scales::breaks_pretty(4), 
  ) +
  facet_wrap(~"Plot 3")

 (p1 + p2) / p3 +
  plot_annotation(
    title = "A collection of plots patched together",
    subtitle = "Works so nice!",
    theme = light_mode_r() +
      theme(
        plot.title = element_text(margin = margin(t = -11)),
        plot.subtitle = element_text(margin = margin(t = 5.5)),
      )
  )
```

### ggtext

The ggtext package support using markdown in text. 

```{r, fig.asp=0.75}
penguins |>
  gg_point(
    x = bill_depth_mm,
    y = bill_length_mm,
    col = species,
    title = "***Pygoscelis*** *penguin* bill **lengths** and depths",
    subtitle = "<span style = 'color: #0095A8 ;'>**Adelie**</span>, 
       <span style = 'color:#FF7043;'>**Chinstrap**</span>, 
       *and* <span style = 'color:#112E51;'>**Gentoo**</span>", 
    mode = light_mode_n()
  ) +
  theme(plot.title = ggtext::element_markdown()) +
  theme(plot.subtitle = ggtext::element_markdown())
```

### showtext

The showtext package enables the use of different fonts in plots. The `*_mode_*` theme functions provide a `base_family` argument.

```{r, fig.asp = 0.7, fig.showtext = TRUE}
head(sysfonts::font_families_google())

sysfonts::font_add_google("Covered By Your Grace", "grace")
sysfonts::font_add_google('Roboto Slab', 'roboto_slab')
sysfonts::font_add_google('Syne Mono', 'syne')

showtext::showtext_auto(enable = TRUE)

penguins |>
  mutate(across(sex, \(x) str_to_sentence(x))) |>
  gg_point(
    x = flipper_length_mm,
    y = body_mass_g,
    col = sex,
    facet = species,
    title = "Penguins body mass by flipper length",
    subtitle = "Palmer Archipelago, Antarctica",
    caption = "Source: Gorman, 2020",
    mode = light_mode_r(base_family = "grace")
  ) +
  theme(
    plot.title = element_text(family = "roboto_slab"),
    plot.subtitle = element_text(family = "syne")
  )

showtext::showtext_auto(enable = FALSE)
```

### ggforce

The ggforce package includes numerous extra geoms, stats etc.

```{r}
library(ggforce)
geom_bspline()

economics |>
  slice_head(n = 35) |> 
  gg_path(
    stat = "bspline", n = 100, type = "clamped",
    x = date, 
    y = unemploy,
    linewidth = 1,
  ) 
```

```{r}
geom_mark_hull()

penguins |>
  gg_blanket(
    geom = "mark_hull",
    x = flipper_length_mm,
    y = body_mass_g,
    col = species,
    coord = coord_cartesian(),
  ) +
  geom_point()
```

### ggbeeswarm

The ggbeeswarm package enables beeswarm plots.

```{r}
library(ggbeeswarm)
geom_beeswarm()
geom_quasirandom()

p1 <- penguins |> 
  gg_point(
    position = position_beeswarm(),
    x = sex, 
    y = flipper_length_mm,
    col = sex,
    x_labels = \(x) str_to_upper(str_sub(x, 1, 1)),
    subtitle = "\nBeeswarm",
    mode = light_mode_n(),
  ) 

p2 <- penguins |> 
  gg_point(
    position = ggbeeswarm::position_quasirandom(),
    x = sex, 
    y = flipper_length_mm,
    col = sex,
    x_labels = \(x) str_to_upper(str_sub(x, 1, 1)),
    subtitle = "\nQuasirandom",
    mode = light_mode_n(),
  ) 

p1 + p2
```

### ggrepel

The ggrepel package can be used to neatly avoid overlapping labels. 

```{r}
library(ggrepel)
geom_text_repel()

mtcars |>
  tibble::rownames_to_column("car") |>
  filter(wt > 2.75, wt < 3.45) |>
  gg_point(
    x = wt, 
    y = mpg, 
    label = car, 
  ) +
  geom_text_repel(colour = blue) 
```

### ggridges

The ggridges package enables ridgeline plots. 

```{r}
library(ggridges)
geom_density_ridges()

ggridges::Catalan_elections |>
  rename_with(snakecase::to_snake_case) |>
  mutate(year = factor(year)) |>
  gg_blanket(
    geom = "density_ridges",
    stat = "density_ridges",
    coord = coord_cartesian(),
    x = percent,
    y = year,
    col = option,
    colour = "white",
    x_limits = c(0, 100),
    y_expand = expansion(c(0, 0.125)),
    alpha = 0.9,
  )
```

### ggdist

The ggdist package enables the visualisation of uncertainty. 

```{r}
library(ggdist)
library(distributional)
geom_slabinterval()

set.seed(123)

data.frame(
  distribution = c("Gamma(2,1)", "Normal(5,1)", "Mixture"),
  values = c(
    dist_gamma(2, 1),
    dist_normal(5, 1),
    dist_mixture(
      dist_gamma(2, 1),
      dist_normal(5, 1),
      weights = c(0.4, 0.6))
  )) |> 
  gg_blanket(
    geom = "slabinterval",
    stat = "slabinterval",
    y = distribution,
    mapping = aes(dist = values),
    colour = blue,
    fill = blue,
    alpha = 0.6,
    y_expand = c(0.05, 0.05),
    mode = light_mode_n(),
    # show_slab = FALSE,
    # show_interval = FALSE,
    # side = "both",
  )
```

```{r}
geom_lineribbon()

set.seed(123)

tibble(x = 1:10) |> 
  group_by(across(everything()))  |> 
  do(tibble(y = rnorm(100, .$x))) |> 
  median_qi(.width = c(0.5, 0.8, 0.95)) |>
  mutate(.width = factor(.width)) |>
  gg_blanket(
    geom = "lineribbon",
    x = x,
    y = y,
    ymin = .lower,
    ymax = .upper,
    col = .width,
    colour = lightness[1],
    col_legend_rev = TRUE,
    col_pal = blues9[c(3,5,7)],
  )
```

### ggdensity

The ggdensity package provides visualizations of density estimates.  

```{r}
library(ggdensity)
geom_hdr()

set.seed(123)

data.frame(
  x = rnorm(1000),
  y = rnorm(1000),
  z = c(rep("A", times = 500), rep("B", times = 500))
) |>
  gg_blanket(
    geom = "hdr", 
    stat = "hdr",
    x = x, 
    y = y, 
    colour = NA,
    y_limits = c(NA, NA),
  ) +
  guides(colour = FALSE, fill = FALSE)
```

```{r}
set.seed(123)

data.frame(
  x = rnorm(1000),
  y = rnorm(1000),
  z = c(rep("A", times = 500), rep("B", times = 500))
) |>
  gg_blanket(
    geom = "hdr", 
    stat = "hdr",
    x = x, 
    y = y, 
    col = z,
    facet = z,
    colour = NA,
    y_limits = c(NA, NA),
  ) 
```

### geomtextpath

The geomtextpath package enables curved text.

```{r}
library(geomtextpath)
geom_textcontour()

expand.grid(x = 1:nrow(volcano), y = 1:ncol(volcano)) |> 
  mutate(z = c(volcano)) |> 
  gg_contour_filled(
    x = x, 
    y = y,
    z = z,
    binwidth = 20,
    x_expand = c(0, 0),
    x_limits = c(NA, NA),
    y_limits = c(NA, NA)) +
  geom_textcontour(
    aes(label = after_stat(level)), 
    size = 2.5, 
    straight = TRUE,
  ) +
  ggeasy::easy_remove_x_axis() + 
  ggeasy::easy_remove_y_axis()
```

### ggblend
 
The ggblend package provides blending of colours. Note you must use a graphics device that supports blending.
 
```{r, fig.asp=0.5}
p1 <- penguins |>
  gg_density(
    x = flipper_length_mm,
    col = species,
    y_limits = c(0, NA),
    mode = light_mode_n(),
  ) +
  facet_wrap(~"Normal") +
  ggeasy::easy_remove_x_axis() +
  ggeasy::easy_remove_y_axis()

p2 <- penguins |>
  gg_blanket(
    stat = "density",
    x = flipper_length_mm,
    col = species,
    y_limits = c(0, NA),
    mode = light_mode_n(),
  ) +
  facet_wrap(~"Blended") +
  geom_density(alpha = 0.6) |> ggblend::blend(blend = "multiply") +
  ggeasy::easy_remove_x_axis() +
  ggeasy::easy_remove_y_axis()

p1 + p2
```

### ggh4x

The ggh4x package includes enhanced facet functionality.

```{r}
iris |> 
  rename_with(\(x) snakecase::to_snake_case(x)) |> 
  mutate(across(species, \(x) str_to_sentence(x))) |> 
  mutate(size = if_else(species == "Setosa", "Short Leaves", "Long Leaves")) |> 
  gg_point(
    x = sepal_width, 
    y = sepal_length, 
    x_breaks = scales::breaks_pretty(n = 3),
  ) +
  ggh4x::facet_nested(
    cols = vars(size, species), 
    nest_line = TRUE, 
    solo_line = TRUE,
  ) +
  theme(strip.text.x = element_text(margin = margin(t = 2.5, b = 5)))
```

### ggeasy

The ggeasy package provides a lot of support for easily modifying themes. Removing axes and gridlines is often useful. 

```{r}
library(ggeasy)

penguins |>
  mutate(across(sex, str_to_sentence)) |> 
  gg_jitter(
    x = species, 
    y = body_mass_g, 
    col = sex, 
  ) +
  light_mode_t() +
  easy_remove_y_gridlines() +
  easy_remove_x_axis(what = c("line", "ticks")) +
  easy_remove_legend_title()
```

### gghighlight

The gghighlight package enables geoms or parts thereof to be highlighted. It _kind of_ works with ggblanket. There must be a `col` variable added unfortunately. It might be better to use ggplot2 instead here.

```{r}
penguins |>
  gg_point(
    x = flipper_length_mm,
    y = body_mass_g,
    col = 1, 
    mode = light_mode_n(),
  ) +
  gghighlight::gghighlight(body_mass_g >= 5000)
```

The `gg_*` function builds the scale for the data that it thinks will be within the panel. Therefore in some situations, you may need to make adjustments, so that the scale builds correctly. 

```{r}
penguins |>
  gg_histogram(
    x = flipper_length_mm,
    col = species,
    x_breaks = scales::breaks_pretty(3),
  ) + #build the scale for not faceted (i.e. all data)
  facet_wrap(~species) + #then add facet layer
  gghighlight::gghighlight() 
```

### ggnewscale

The ggnewscale package enables multiple colour scales.

```{r}
penguins |> 
  gg_blanket(
    x = flipper_length_mm, 
    y = body_mass_g,
  ) + 
  geom_point(
    mapping = aes(colour = bill_length_mm), 
    data = penguins |> filter(species == "Gentoo"),
  ) +
  scale_color_viridis_c(option = "G", direction = -1) +
  labs(colour = "Gentoo\nBill length (mm)") +
  ggnewscale::new_scale_colour() +
  geom_point(
    mapping = aes(colour = species), 
    data = penguins |> filter(species != "Gentoo"),
  ) +
  scale_color_manual(values = c("#901752", orange)) +
  labs(colour = "Species")
```


