---
title: "Go further"
author: "David Hodge"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.asp = 0.6,
  out.width = "70%",
  dpi = 300
  )
```

## Overview

This article will demonstrate a random assortment of content, including some of which is more advanced.  

* Change the `stat` of the layer
* Change the `position` of the layer
* Reorder and/or reverse categorical variables
* Drop unused categorical variable values
* Transform scales to `"log"` etc
* Correct the default aspect
* Avoid 'symmetric' scales
* Change the `*_position` of positional axes 
* Zoom in or out on scales
* Use delayed evaluation
* Rescale a diverging col scale
* Add a legend within the panel
* Set consistent panel sizes
* Set consistent widths across panels
* Multiple geoms

```{r setup}
library(dplyr)
library(tidyr)
library(forcats)
library(stringr)
library(ggplot2)
library(scales)
library(ggblanket)
library(patchwork)
library(palmerpenguins)

set_blanket()

penguins2 <- penguins |> 
  labelled::set_variable_labels(
    bill_length_mm = "Bill length (mm)",
    bill_depth_mm = "Bill depth (mm)",
    flipper_length_mm = "Flipper length (mm)",
    body_mass_g = "Body mass (g)",
  ) |> 
  mutate(sex = stringr::str_to_sentence(sex)) 
```

### Change the `stat` of the layer

The default `stat` of each `gg_*` function can be changed. 

```{r}
penguins2 |>
  gg_pointrange(
    stat = "summary", 
    x = species,
    y = flipper_length_mm, 
  )
```

```{r}
library(ggforce)

ggplot2::economics |>
  slice_head(n = 35) |> 
  gg_path(
    stat = "bspline", n = 100,
    x = date, 
    y = unemploy,
    y_title = "Unemployment",
  ) 
```

### Change the `position` of the layer

The default `position` of each `gg_*` function can be changed. 

```{r}
penguins2 |> 
  gg_point(
    position = ggbeeswarm::position_quasirandom(),
    x = sex, 
    y = flipper_length_mm,
    col = sex,
  ) +
  theme(legend.position = "none")
```


### Reorder and/or reverse categorical variables

ggblanket requires unquoted variables only for `x`, `y`, `col`, `facet`, `facet2` and  `alpha`. You can often manipulate the data prior to plotting to achieve what you want (e.g. using `tidyr::drop_na`, `forcats::fct_rev` and/or `forcats::fct_reorder`).

```{r}
p1 <- diamonds |>
  count(color) |>
  gg_col(
    x = n,
    y = color,
    width = 0.75,
    x_labels = \(x) x / 1000,
    x_title = "Count (thousands)", 
    subtitle = "\nDefault order"
  )

p2 <- diamonds |>
  count(color) |>
  mutate(color = fct_rev(fct_reorder(color, n))) |>
  gg_col(
    x = n,
    y = color,
    width = 0.75,
    x_labels = \(x) x / 1000,
    x_title = "Count (thousands)", 
    subtitle = "\nRe-orderered"
  ) 

p1 + p2
```

### Drop unused categorical variable values

ggblanket keeps unused factor levels in the plot. If users wish to drop unused levels they should likewise do it in the data prior to plotting using `forcats::fct_drop`. 

```{r}
p1 <- diamonds |> 
  count(color) |>
  filter(color %in% c("E", "G", "I")) |>
  gg_col(
    x = n,
    y = color,
    width = 0.75,
    x_labels = \(x) x / 1000,
    x_title = "Count (thousands)", 
    subtitle = "\nUnused levels kept",
  )

p2 <- diamonds |> 
  count(color) |>
  filter(color %in% c("E", "G", "I")) |>
  mutate(color = forcats::fct_drop(color)) |> 
  gg_col(
    x = n,
    y = color,
    width = 0.75,
    x_labels = \(x) x / 1000,
    x_title = "Count (thousands)", 
    subtitle = "\nUnused levels dropped",
  )

p1 + p2
```

### Transform scales to `"log"` etc

Transform objects (e.g. `transform_log()` or character strings of these can be used to transform scales - including combining these.

```{r, fig.asp=0.75}
p1 <- pressure |>
  gg_point(
    x = temperature, 
    y = pressure, 
    x_breaks_n = 4,
    y_breaks_n = 4,
    subtitle = "\nDefault",
  )

p2 <- pressure |>
  gg_point(
    x = temperature, 
    y = pressure, 
    x_breaks_n = 4,
    y_breaks_n = 4,
    y_transform = "reverse", 
    subtitle = "\nReverse",
  )

p3 <- pressure |>
  gg_point(
    x = temperature, 
    y = pressure,
    x_breaks_n = 4,
    y_breaks_n = 4,
    y_transform = "sqrt",
    subtitle = "\nSqrt", 
  )

p4 <- pressure |>
  gg_point(
    x = temperature, 
    y = pressure, 
    x_breaks_n = 4,
    y_breaks_n = 4,
    y_transform = c("sqrt", "reverse"),
    subtitle = "\nSqrt & Reverse",
  )

(p1 + p2) / (p3 + p4)
```

### Correct the default aspect (and symmetric scales)

The `gg_*` function guesses the `aspect` of the plot to determine what side-effects to have on the set theme. If it guesses incorrectly, use the `aspect` argument.

```{r}
p1 <- palmerpenguins::penguins |>
  gg_histogram(
    y = flipper_length_mm,
    col = sex,
    bins = 50,
    subtitle = "\nDefault aspect  (and symmetric scales)",
  )

p2 <- palmerpenguins::penguins |>
  gg_histogram(
    y = flipper_length_mm,
    col = sex,
    bins = 50,
    aspect = "y",
    subtitle = "\nDefault aspect (and symmetric scales)",
  )

p1 + p2
```

### Avoid 'symmetric' scales

Symmetric scales can be turned off or on using the `symmetric` argument. 

```{r}
p1 <- penguins2 |>
  gg_pointrange(
    x = sex,
    y = bill_length_mm,
    stat = "summary",
    position = position_dodge(),
    x_labels = \(x) str_sub(x, 1, 1),
    subtitle = "\nsymmetric = 'y'",
  ) +
  labs(y = NULL)

p2 <- penguins2 |>
  gg_pointrange(
    x = sex,
    y = bill_length_mm,
    stat = "summary",
    position = position_dodge(),
    symmetric = "none",
    x_labels = \(x) str_sub(x, 1, 1),
    subtitle = "\nsymmetric = 'none'",
  ) +
  labs(y = NULL)

p1 + p2
```

### Change the `*_position` of positional axes 

Positional axes can be changed using `*_position`.

Note that for `x_position = "top"`, a caption must be added or modified to make this work nicely with a `*_mode_*` theme. 

```{r}
economics |>
  gg_line(
    x = date,
    y = unemploy,
    col = date,
    y_position = "right",
    x_position = "top",
    caption = "", 
    title = "Unemployment",
    subtitle = "1967\u20132015",
  ) 
```

### Zoom in or out on scales

There are no `*_limits` arguments in ggblanket. 

Instead, users should use a combination of filtering the data, adding `*_expand_limits` and `coord = coord_cartesian(xlim = ..., ylim = ...)` arguments etc. 

```{r}
#To Zoom out, use *_expand_limits:
penguins |>
  gg_smooth(
    x =  body_mass_g, 
    y = bill_depth_mm,
    x_limits_include = c(0),
    y_limits_include = c(10, 25),
    se = TRUE,
  ) 
```
```{r}
#To zoom-in when the stat equals "identity", use dplyr::filter
penguins |>
  filter(bill_depth_mm < 15) |>
  gg_point(
    x =  bill_depth_mm, 
    y = body_mass_g,
  ) 
```

```{r}
#To zoom-in when the stat does _not_ equal "identity", use coord_cartesian 
#Then either recreate the breaks, or turn off the symmetric axis 
penguins |>
  gg_smooth(
    x =  body_mass_g, 
    y = bill_depth_mm,
    coord = coord_cartesian(ylim = c(14.8, 15)),
    y_breaks = scales::breaks_width(0.05),
    se = TRUE,
  ) 
```

###  Use delayed evaluation

 The `mapping` argument can be used for delayed evaluation with the `ggplot2::after_stat` function.

```{r, fig.asp=0.55}
penguins2 |>
  gg_histogram(
    x = flipper_length_mm,
    mapping = aes(y = after_stat(density)),
    facet = species,
  )
```

```{r, fig.asp=0.55}
faithfuld |>
  gg_contour(
    x = waiting,
    y = eruptions,
    z = density,
    mapping = aes(colour = after_stat(level)),
    bins = 8,
  )
```

### Rescale a diverging col scale

Use `col_rescale` to rescale a diverging scale around a central point. 

```{r}
rescale_vctr <- sort(c(range(mpg$cty), 15))

mpg  |>  
  gg_point(
    x = displ, 
    y = hwy, 
    col = cty,
    col_rescale = scales::rescale(rescale_vctr),
    col_breaks = scales::breaks_width(5),
    col_palette = c(jumble[c(3, 1)], "white", jumble[c(2, 4)]),
  )
```

### Add a legend within the panel

```{r}
set_blanket()

penguins2 |> 
  gg_histogram(
    x = flipper_length_mm,
    col = species,
  ) +
  theme(legend.position = "inside") +
  theme(legend.position.inside = c(1, 0.975)) +
  theme(legend.justification = c(1, 1)) 
```

### Set consistent panel sizes

Use `panel_heights` and `panel_widths` arguments in `theme_lighter`/`theme_darker` - or `+ theme(panel.heights = NULL, panel.widths = NULL)`. Use `rep` to make them all the same.

```{r}
set_blanket(
  theme = theme_lighter(
    panel_heights = rep(unit(50, "mm"), times = 100), 
    panel_widths = rep(unit(50, "mm"), times = 100),
  )
)

penguins |>
  gg_point(
    x = bill_length_mm,
    y = body_mass_g,
    col = species,
    size = 1,
  )

penguins |>
  drop_na(sex) |> 
  mutate(across(sex, \(x) str_to_sentence(x))) |> 
  gg_point(
    x = bill_length_mm,
    y = body_mass_g,
    col = species,
    facet = sex,
    size = 1,
  )
```

### Set consistent widths across panels

The width of geoms can be standardised across geoms using `standardise_width`. 

```{r}
set_blanket(
  theme = theme_lighter(
    panel_heights = rep(unit(50, "mm"), times = 100),
    panel_widths = rep(unit(75, "mm"), times = 100),
  )
)

# create standard
penguins |>
  drop_na(sex) |>
  mutate(across(sex, \(x) str_to_sentence(x))) |>
  gg_bar(
    x = species,
    col = sex,
    width = 0.25,
  )

# align width with standard
penguins |>
  drop_na(sex) |>
  mutate(across(sex, \(x) str_to_sentence(x))) |>
  gg_bar(
    x = sex,
    col = species,
    width = standardise_width(from_n = 2, to_width = 0.25, to_n = 3)
  )

penguins |>
  drop_na(sex) |>
  mutate(across(sex, \(x) str_to_sentence(x))) |>
  gg_bar(
    x = sex,
    col = species,
    position = position_dodge(),
    width = standardise_width(from_n = 2, from_dodge_n = 3, to_width = 0.25, to_n = 3)
  )

penguins |>
  drop_na(sex) |>
  mutate(across(sex, \(x) str_to_sentence(x))) |>
  gg_bar(
    y = sex,
    col = species,
    position = position_dodge(),
    width = standardise_width(from_n = 2, 
                              from_dodge_n = 3,
                              from_aspect = "y",
                              to_width = 0.25, 
                              to_n = 3)
  )

set_blanket(
  theme = theme_lighter(
    panel_heights = rep(unit(40, "mm"), times = 100),
    panel_widths = rep(unit(100, "mm"), times = 100),
  )
)
```

### Annotate an axis line

An axis line can be easily replaced by a new annotated axis line, which can sometimes be useful. This function only works where the panel heights and panel widths are set in the theme.

```{r}
set_blanket(
  theme = theme_lighter(
    panel_heights = rep(unit(50, "mm"), times = 100),
    panel_widths = rep(unit(75, "mm"), times = 100),
  ),
)

penguins |>
  tidyr::drop_na(sex) |>
  mutate(across(sex, \(x) str_to_sentence(x))) |>
  add_row(flipper_length_mm = 195, body_mass_g = 2500, sex = "Female") |> 
  gg_blanket(
    x = flipper_length_mm,
    y = body_mass_g,
    col = sex,
  ) +
  annotate_axis_line() +
  geom_point()
```

### Annotate axis ticks

Likewise, axis ticks can be easily be replaced by new annotated axis ticks, which is sometimes be useful.

 This function only works where the panel heights and panel widths are set in the theme.

```{r}
set_blanket(
  theme = theme_lighter(
    panel_heights = rep(unit(50, "mm"), times = 100),
    panel_widths = rep(unit(75, "mm"), times = 100),
  ),
)

penguins |>
  tidyr::drop_na(sex) |>
  mutate(across(sex, \(x) str_to_sentence(x))) |>
  gg_histogram(
    x = flipper_length_mm,
    col = sex,
  ) +
  annotate_axis_ticks(x_breaks = c(seq(170, 230, 10), 195))
```

### Multiple geoms

ggblanket is primarily build for single geoms. However, users can add multiple `ggplot2::geom_*` layers. 

The `gg_*()` geom layer will be the bottom geom layer of the plot, and each subsequent `geom_*()` layer is placed on top. 

```{r}
palmerpenguins::penguins |> 
  gg_violin(
    x = species, 
    y = bill_depth_mm,
    outliers = FALSE,
  ) +
  geom_jitter(shape = 19, colour = col_multiply(jumble[11])) 
```

The scales are built within the `gg_*()` function _without_ knowledge of later layers. The `gg_*()` function builds scales with regard to the `stat`, `position`, and aesthetics (that the geom understands) etc. So, in some situations, users will need to take care. 

```{r}
palmerpenguins::penguins |>
  drop_na(sex) |> 
  group_by(species, sex) |>
  summarise(
    lower = quantile(bill_depth_mm, probs = 0.05),
    upper = quantile(bill_depth_mm, probs = 0.95),
    bill_depth_mm = mean(bill_depth_mm, na.rm = TRUE),
  ) |>
  labelled::copy_labels_from(palmerpenguins::penguins) |>
  gg_blanket(
    y = species,
    x = bill_depth_mm,
    xmin = lower, 
    xmax = upper,
    col = sex,
    bordered = TRUE,
    x_limits_include = 0,
  ) +
  geom_col(
    width = 0.75,
    position = position_dodge(),
  ) +
  geom_errorbar(
    width = 0.1, 
    position = position_dodge(width = 0.75),
  ) 
```
