---
title: "Extending further"
author: "David Hodge"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.asp = 0.618,
  out.width = "70%",
  dpi = 300
)
```

## Overview

To extend ggblanket further, users can:

1. Join plots together using the patchwork package
2. Make plots that use the ggtext package
3. Make plots with non-supported geoms
4. Make plots with non-supported aesthetics
5. Use non-supported arguments within ggplot2 layers
6. Make plots with multiple geom layers
7. Create custom functions that wrap ggblanket functions

```{r setup}
library(dplyr)
library(ggplot2)
library(ggblanket)
library(palmerpenguins)
library(patchwork)

penguins <- penguins |>
  mutate(sex = stringr::str_to_sentence(sex)) |>
  tidyr::drop_na(sex)
```

### 1. Join plots together using the patchwork package

The fantastic patchwork package can be used to patch plots together.

```{r}
p1 <- penguins |> 
  gg_point(
    x = flipper_length_mm, 
    y = body_mass_g, 
    col = species, 
    x_breaks = scales::breaks_pretty(3), 
    col_legend_ncol = 2) 

p2 <- penguins |> 
  gg_jitter(
    x = species, 
    y = body_mass_g, 
    col = sex, 
    pal = c("#2B6999", "#E37000"),
    y_labels = \(x) stringr::str_sub(x, 0, 0),
    y_title = "")

p1 + p2
```

### 2. Make plots that use the ggtext package

You can modify plot text using the ggtext package. For titles, it's important to make the theme have a plain title first before adding a theme layer to state that the title is in markdown. 

```{r}
nc <- sf::st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)

nc |>
  gg_sf(col = AREA,
        pal = RColorBrewer::brewer.pal(9, "Reds"),
        title = "**Bold** or _italics_ or <span style = 'color:red;'>red</span>",
        theme = gg_theme(title_face = "plain")) +
  theme(plot.title = ggtext::element_markdown())
```

### 3. Make plots with non-supported geoms

There are lots of non-supported geoms in ggplot2 and other extension packages that can be used in combination with   ggblanket. 

```{r}
iris |>
  mutate(Species = stringr::str_to_sentence(Species)) |> 
  gg_blank(
    x = Sepal.Width,
    y = Sepal.Length,
    col = Species,
    facet = Species,
    col_legend_place = "r") +
  ggdensity::geom_hdr(colour = NA) +
  labs(alpha = "Probs") +
  theme(legend.title = element_text(margin = margin(t = 10)))
```

### 4. Make plots with non-supported aesthetics

While aesthetics other than col/fill are not supported by ggblanket, it is possible to access these in plots made with a combination of `gg_blank` and other ggplot2 code.

```{r}
#for discrete non-col aesthetic & col aesthetic
penguins |>
  gg_blank(
    x = flipper_length_mm,
    y = body_mass_g, 
    col = sex) + 
  geom_point(aes(shape = sex)) +  
  labs(shape = "Sex") 
```

```{r}
#for discrete non-col aesthetic & no col aesthetic
penguins |>
  gg_blank(
    x = flipper_length_mm,
    y = body_mass_g, 
    col = sex) + 
  geom_point(aes(shape = sex), col = pal_default(1)) + 
  labs(shape = "Sex") 
```

```{r}
#for continuous non-col aesthetic & col aesthetic
penguins |>
  gg_blank(
    x = flipper_length_mm,
    y = body_mass_g, 
    col = species,
    col_legend_place = "r") + 
  geom_point(aes(size = bill_depth_mm, col = species)) +
  labs(size = "Bill depth mm") +
  theme(legend.title = element_text(margin = margin(t = 10)))
```

```{r}
#for continuous non-col aesthetic & no col aesthetic
penguins |>
  gg_blank(
    x = flipper_length_mm,
    y = body_mass_g) + 
  geom_point(aes(size = bill_depth_mm), col = pal_default(1)) +
  labs(size = "Bill depth mm") 
```

### 5. Use non-supported arguments within ggplot2 layers

The switch argument within facet_grid is not supported within ggblanket. As the `...` relates to the `geom_*` function, this cannot be accessed. A way to access this functionality is to `+` the `facet_grid(..., switch = "y")` in.

```{r}
penguins |>
  gg_point(
    x = flipper_length_mm, 
    y = body_mass_g,
    col = island,
    facet = species, 
    facet2 = sex
  ) +
  facet_grid(cols = vars(species),
             rows = vars(sex), 
             switch = "y") +
  theme(strip.switch.pad.grid = unit(2, "mm"))
```

### 6. Make plots with multiple geom layers

The `gg_*` function puts the aesthetics within the wrapped `ggplot` function. Therefore, the aesthetics will inherit to any subsequent geom's added by default. Geom's will plot in order. The `gg_*` function will plot the associated geom as the first layer, and then other geoms will be plotted on top of it. Use the `*_include` and `*_limits` arguments as necessary to ensure the min and max of all layers are included in your numeric scales.

```{r}
penguins |>
  gg_boxplot(x = species,
             y = body_mass_g,
             width = 0.5,
             pal = "#1B9E77", 
             outlier.colour = NA) +
  geom_jitter(col = "#1B9E77")
```

If some geoms have a `col` aesthetic and some do not, then it is useful to have the col aesthetic in the `gg_*` function. If you do not actually want a col aesthetic in this layer, then you can over-rule the aesthetic by adding colour and fill arguments. In some circumstances, it may be useful to use `gg_blank`.

```{r}
penguins |>
  group_by(sex, species) |>
  summarise(across(body_mass_g, mean)) |>
  mutate(upper = body_mass_g * 1.05) |>
  mutate(lower = body_mass_g * 0.95) %>%
  gg_col(
    x = sex,
    y = body_mass_g,
    col = sex,
    facet = species,
    y_include = max(.$upper),
    fill = "#d3d3d3",
    colour = "#d3d3d3") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.1)
```

### 7. Create custom functions that wrap ggblanket functions

You can easily create powerful custom functions. This is because the `...` argument can allow you to access _all_  arguments within the ggblanket `gg_*` function (and applicable `ggplot2::geom_*` function). 

```{r}
gg_point_custom <- function(data, x, y, col, 
                            size = 3, 
                            shape = 17,
                            pal = pals::brewer.dark2(9), 
                            col_title = "", 
                            col_legend_place = "t",
                            ...) {
  data |> 
    gg_point(x = {{ x }}, y = {{ y }}, col = {{col}}, 
             size = size, 
             shape = shape,
             pal = pal, 
             col_title = col_title, 
             col_legend_place = col_legend_place, 
             col_labels = stringr::str_to_sentence,
             ...)
}

iris |>
  gg_point_custom(
    x = Sepal.Width,
    y = Sepal.Length,
    col = Species)
```

```{r}
penguins |>
  gg_point_custom(
    x = species,
    y = body_mass_g,
    col = sex,
    position = position_jitter())
```

